### 2017-06-28 inst/doc/msde-quicktut.Rmd
added inst/doc to repo

---
title: "Inference for multivariate Stochastic Differential Equations with **`msde`**"
author: "Martin Lysy, JunYong Tong, Nigel Delaney"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{msde-quicktut}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\tth}{\bm{\theta}}
\newcommand{\pphi}{\bm{\phi}}
\newcommand{\eeta}{\bm{\eta}}
\newcommand{\Y}{\bm{Y}}
\newcommand{\y}{\bm{y}}
\newcommand{\B}{\bm{B}}
\renewcommand{\a}{\alpha}
\renewcommand{\b}{\beta}
\newcommand{\g}{\gamma}
\newcommand{\rv}[3][1]{#2_{#1},\ldots,#2_{#3}}
\newcommand{\dr}{\bm{\Lambda}}
\newcommand{\df}{\bm{\Sigma}}
\newcommand{\dt}{\Delta t}
\newcommand{\ud}{\mathrm{d}}
\newcommand{\var}{\mathrm{var}}
\newcommand{\cov}{\mathrm{cov}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\Ym}{\Y_{\mathrm{miss}}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\ind}{\stackrel{\mathrm{ind}}{\sim}}

## Introduction

A $d$-dimensional stochastic differential equation (SDE) $\Y_t = (\rv [1t] Y {dt})$ is written as
$$
\ud \Y_t = \dr_{\tth}(\Y_t)\,\ud t + \df_{\tth}(\Y_t)^{1/2}\,\ud \B_t,
$$
where $\dr_{\tth}(\y)$ and $\df_{\tth}(\y)$ are the drift and diffusion functions,
<!-- $$ -->
<!-- \dr_{\tth}(\y) = \lim_{\dt \to 0} \frac{E[\Y_{t+\dt} - \Y_t \mid \Y_t = \y]}{\dt}, \qquad \df_{\tth}(\y) = \lim_{\dt \to 0} \frac{\var(\Y_{t+\dt} - \Y_t \mid \Y_t = \y)}{\dt}, -->
<!-- $$ -->
and $\B_t = (\rv [1t] B {dt})$ is $d$-dimensional Brownian motion.  The **`msde`** package implements a Markov Chain Monte Carlo (MCMC) algorithm to sample from the posterior distribution $p(\tth \mid \Y)$ of the parameters given discrete observations $\Y = (\rv [0] {\Y} N)$ recorded at times $\rv [0] t N$, with some of the $d$ components of $\Y_t$ possibly latent.  To do this efficiently, **`msde`** requires on-the-fly **C++** compiling of user-specified models.  Instructions for setting up **R** to compile **C++** code are provided in the [Installation](#install) section.

## Creating an `sde.model` object

The SDE model used throughout this vignette is the so-called Lotka-Volterra predator-prey model.  Let $H_t$ and $L_t$ denote the number of Hare and Lynx at time $t$ coexisting in a given habitat.  The Lotka-Volterra SDE describing the interactions between these two animal populations is given by [@golightly-wilkinson10]:
$$
\begin{pmatrix} \mathrm{d} H_t \\ \mathrm{d} L_t \end{pmatrix} = \begin{pmatrix} \a H_t - \b H_tL_t \\ \b H_tL_t - \g L_t \end{pmatrix}\, \mathrm{d} t + \begin{pmatrix} \a H_t + \b H_tL_t & -\b H_tL_t \\ -\b H_tL_t & \b H_tL_t + \g L_t\end{pmatrix}^{1/2} \begin{pmatrix} \mathrm{d} B_{1t} \\ \mathrm{d} B_{2t} \end{pmatrix}.
$$
Thus we have $d = 2$, $\Y_t = (H_t, L_t)$, and $\tth = (\a, \b, \g)$.

### The `sdeModel` class definition

In order to build this model in **C++**, we create a header file `lotvolModel.h` containing the class definition for an `sdeModel` object.  The basic structure of this class is given below:
```{Rcpp, eval = FALSE}
// sde model object
class sdeModel {
 public:
  static const int nParams = 3; // number of model parameters
  static const int nDims = 2; // number of sde dimensions
  static const bool sdDiff = false; // whether diffusion function is on sd or var scale
  static const bool diagDiff = false; // whether diffusion function is diagonal
  void sdeDr(double *dr, double *x, double *theta); // drift function
  void sdeDf(double *df, double *x, double *theta); // diffusion function
  bool isValidParams(double *theta); // parameter validator
  bool isValidData(double *x, double *theta); // data validator
};
```
The meaning of each class member is as follows:

* `nParams`: The number of model parameters.  For the Lotka-Volterra model we have $\tth = (\a, \b, \g)$, such that `nParams = 3`.
* `nDims`: The number of dimensions in the multivariate SDE.  In this case we have $\Y_t = (H_t, L_t)$, such that `nDims = 2`.
* `sdeDr`: The SDE drift function.  In **R**, this function would be implemented as
    ```{r, eval = FALSE}
sde.drift <- function(x, theta) {
  dr <- c(theta[1]*x[1] - theta[2]*x[1]*x[2], # alpha * H - beta * H*L
          theta[2]*x[1]*x[2] - theta[3]*x[2]) # beta * H*L - gamma * L
  dr
}
```
    In **C++** the same thing is accomplished with
    ```{Rcpp, eval = FALSE}
void sdeDr(double *dr, double *x, double *theta) {
  dr[0] = theta[0]*x[0] - theta[1]*x[0]*x[1]; // alpha * H - beta * H * L
  dr[1] = theta[1]*x[0]*x[1] - theta[2]*x[1]; // beta * H * L - gamma * L
  return;
}
```
* `sdeDf`: The SDE diffusion function.  This can be specified on the standard deviation scale, or on the variance scale as above.  In this case, an **R** implementation would be
    ```{r, eval = FALSE}
sde.diff <- function(x, theta) {
  df <- matrix(NA, 2, 2)
  df[1,1] <- theta[1]*x[1] + theta[2]*x[1]*x[2] # alpha * H + beta * H*L
  df[1,2] <- -theta[2]*x[1]*x[2] # -beta * H*L
  df[2,1] <- df[1,2] # -beta * H*L
  df[2,2] <- theta[2]*x[1]*x[2] + theta[3]*x[2] # beta * H*L + gamma * L
  df
}
```
    In **C++** the specification is slightly different.  First we set `sdDiff = false` in order to tell **`msde`** to use the variance scale.  Next, the diffusion function is coded as
    ```{Rcpp, eval = FALSE}
void sdeDf(double *df, double *x, double *theta) {
  df[0] = theta[0]*x[0] + theta[1]*x[0]*x[1]; // matrix element (1,1)
  df[2] = -theta[1]*x[0]*x[1]; // element (1,2)
  df[3] = theta[1]*x[0]*x[1] + theta[2]*x[1]; // element (2,2)
  return;
}
```
    Thus there are two major differences with the **R** version.  The first is that the `df` matrix is stored as a vector (or "array" in **C++**).  Its elements are stored in column-major order, i.e., by stacking the columns one after the other into one long vector.  The second difference is that **`msde`** only uses the *upper triangular* portion of the (symmetric) matrix $\df_{\tth}(\y)$.  The elements below the diagonal can be set to any value or not set at all without affecting the computations.

	**`msde`** internally computes the Cholesky decomposition of the diffusion function when it is specified on the variance scale.  For small problems such as this one, it is more efficient to specify the Cholesky decomposition directly, i.e., specify the diffusion on the standard deviation scale.  In this case, the Cholesky decomposition of the diffusion function is
$$
\begin{pmatrix} \a H_t + \b H_tL_t & -\b H_tL_t \\ -\b H_tL_t & \b H_tL_t + \g L_t\end{pmatrix}^{1/2} = \begin{pmatrix} \sqrt{\a H_t + \b H_tL_t} & -\frac{\b H_tL_t}{\sqrt{\a H_t + \b H_tL_t}} \\ 0 & \sqrt{\b H_tL_t + \g L_t - \frac{(\b H_tL_t)^2}{\a H_t + \b H_tL_t}} \end{pmatrix},
$$
which is passed to **`msde`** by setting `sdDiff = true` and
    ```{Rcpp, eval = FALSE}
void sdeDf(double *df, double *x, double *theta) {
  double bHL = theta[1]*x[0]*x[1]; // beta * H*L
  df[0] = sqrt(theta[0]*x[0] + bHL); // sqrt(alpha * H + bHL)
  df[2] = -bHL/df[0];
  df[3] = sqrt(bHL + theta[2]*x[1] - df[2]*df[2]);
  return;
}
```
* `diagDiff`: A logical specifying whether or not the diffusion matrix $\df_\tth(\y)$ is diagonal.  For the Lotka-Volterra SDE model above, we set `diagDiff = false`.  However, if the diffusion function were of the form
$$
\df_\tth(\Y_t) = \begin{pmatrix} \a H_t + \b H_tL_t & 0 \\ 0 & \b H_tL_t + \g L_t\end{pmatrix},
$$
then we would set `diagDiff = true`, **and importantly**, treat the `df` argument of `sdeDf` directly as the diagonal elements of the diffusion matrix.  That is, the diffusion (on the variance scale) is encoded as
    ```{Rcpp, eval = FALSE}
void sdeDf(double *df, double *x, double *theta) {
  double bHL = theta[1]*x[0]*x[1]; // beta * H*L
  df[0] = theta[0]*x[0] + bHL; // alpha * H + bHL
  // note the assignment to df[1] and not df[3]
  df[1] = bHL + theta[2]*x[1]; // bHL + gamma * L
  return;
}
```
* `isValidParams`: A logical used to specify the parameter support.  In this case we have $\a, \b, \g > 0$, such that
    ```{Rcpp, eval = FALSE}
bool isValidParams(double *theta) {
  bool val = theta[0] > 0.0;
  val = val && theta[1] > 0.0;
  val = val && theta[2] > 0.0;
  return val;
}
```
* `isValidData`: A logical used to specify the SDE support, which can be parameter-dependent.  In this case we simply have $H_t, L_t > 0$, such that
    ```{Rcpp, eval = FALSE}
bool isValidData(double *x, double *theta) {
  return (x[0] > 0.0) && (x[1] > 0.0);
}
```

Thus the whole file `lotvolModel.h` is given below:
```{Rcpp, eval = FALSE}
#ifndef sdeModel_h
#define sdeModel_h 1

// Lotka-Volterra Predator-Prey model

// class definition
class sdeModel {
 public:
  static const int nParams = 3; // number of model parameters
  static const int nDims = 2; // number of sde dimensions
  static const bool diagDiff = false; // whether diffusion function is diagonal
  static const bool sdDiff = true; // whether diffusion is on sd or var scale
  void sdeDr(double *dr, double *x, double *theta); // drift function
  void sdeDf(double *df, double *x, double *theta); // diffusion function
  bool isValidParams(double *theta); // parameter validator
  bool isValidData(double *x, double *theta); // data validator
};

// drift function
inline void sdeModel::sdeDr(double *dr, double *x, double *theta) {
  dr[0] = theta[0]*x[0] - theta[1]*x[0]*x[1]; // alpha * H - beta * H*L
  dr[1] = theta[1]*x[0]*x[1] - theta[2]*x[1]; // beta * H*L - gamma * L
  return;
}

// diffusion function (sd scale)
inline void sdeModel::sdeDf(double *df, double *x, double *theta) {
  double bHL = theta[1]*x[0]*x[1]; // beta * H*L
  df[0] = sqrt(theta[0]*x[0] + bHL); // sqrt(alpha * H + bHL)
  df[2] = -bHL/df[0];
  df[3] = sqrt(bHL + theta[2]*x[1] - df[2]*df[2]);
  return;
}

// parameter validator
inline bool sdeModel::isValidParams(double *theta) {
  bool val = theta[0] > 0.0;
  val = val && theta[1] > 0.0;
  val = val && theta[2] > 0.0;
  return val;
}

// data validator
inline bool sdeModel::isValidData(double *x, double *theta) {
  return (x[0] > 0.0) && (x[1] > 0.0);
}

#endif
```
The additions to the previous code sections are:

1. The header include guards (`#ifndef`/`#define`/`#endif`).
2. The `sdeModel::` identifier is prepended to the class member definitions when these are written outside of the class declaration itself.
3. The `inline` keyword before the class member definitions, both of which ensure that only one instance of these functions is passed to the **C++** compiler.

### Compiling and checking the `sde.model` object

One the `sdeModel` class is created as the **C++** level, it is compiled in **R** using the following commands:
```{r}
require(msdeHeaders)

# put lotvolModel.h in the working directory
data.names <- c("H", "L")
param.names <- c("alpha", "beta", "gamma")
lvmod <- sde.make.model(ModelFile = "lotvolModel.h",
                        data.names = data.names,
                        param.names = param.names)
```
Before using the model for inference, it is useful to make sure that the **C++** entrypoints are error-free.  To facilitate this, **`msde`** provides **R** wrappers to the internal **C++** drift, diffusion, and validator functions, which can then be checked against **R** versions as follows:
```{r}
# helper functions

# random matrix of size nreps x length(x) from vector x
jit.vec <- function(x, nreps) {
  apply(t(replicate(n = nreps, expr = x, simplify = "matrix")), 2, jitter)
}
# maximum absolute and relative error between two arrays
max.diff <- function(x1, x2) {
  c(abs = max(abs(x1-x2)), rel = max(abs(x1-x2)/max(abs(x1), 1e-8)))
}

# R sde functions

# drift and diffusion
lv.drift <- function(x, theta) {
  dr <- c(theta[1]*x[1] - theta[2]*x[1]*x[2], # alpha * H - beta * H*L
          theta[2]*x[1]*x[2] - theta[3]*x[2]) # beta * H*L - gamma * L
  dr
}
lv.diff <- function(x, theta) {
  df <- matrix(NA, 2, 2)
  df[1,1] <- theta[1]*x[1] + theta[2]*x[1]*x[2] # alpha * H + beta * H*L
  df[1,2] <- -theta[2]*x[1]*x[2] # -beta * H*L
  df[2,1] <- df[1,2] # -beta * H*L
  df[2,2] <- theta[2]*x[1]*x[2] + theta[3]*x[2] # beta * H*L + gamma * L
  chol(df) # always use sd scale in R
}

# validators
lv.valid.data <- function(x, theta) all(x > 0)
lv.valid.params <- function(theta) all(theta > 0)

# generate some test values
nreps <- 12
x0 <- c(H = 71, L = 79)
theta0 <- c(alpha = .5, beta = .0025, gamma = .3)
X <- jit.vec(x0, nreps)
Theta <- jit.vec(theta0, nreps)

# drift and diffusion check

# R versions
dr.R <- matrix(NA, nreps, lvmod$ndims) # drift
df.R <- matrix(NA, nreps, lvmod$ndims^2) # diffusion
for(ii in 1:nreps) {
  dr.R[ii,] <- lv.drift(x = X[ii,], theta = Theta[ii,])
  # flattens diffusion matrix into a row
  df.R[ii,] <- c(lv.diff(x = X[ii,], theta = Theta[ii,]))
}

# C++ versions
dr.cpp <- sde.drift(model = lvmod, x = X, theta = Theta)
df.cpp <- sde.diff(model = lvmod, x = X, theta = Theta)

# compare
max.diff(dr.R, dr.cpp)
max.diff(df.R, df.cpp)

# validator check

# generate invalid data and parameters
X.bad <- X
X.bad[c(1,3,5),1] <- -X.bad[c(1,3,5),1]
Theta.bad <- Theta
Theta.bad[c(2,4,6),3] <- -Theta.bad[c(2,4,6),3]

# R versions
x.R <- rep(NA, nreps)
theta.R <- rep(NA, nreps)
for(ii in 1:nreps) {
  x.R[ii] <- lv.valid.data(x = X.bad[ii,], theta = Theta.bad[ii,])
  theta.R[ii] <- lv.valid.params(theta = Theta.bad[ii,])
}

# C++ versions
x.cpp <- sde.valid.data(model = lvmod, x = X.bad, theta = Theta.bad)
theta.cpp <- sde.valid.params(model = lvmod, theta = Theta.bad)

# compare
c(x = all(x.R == x.cpp), theta = all(theta.R == theta.cpp))

```

## Simulating trajectories from the Lotka-Volterra model

The basis for both simulation and inference with SDEs is the Euler-Maruyama approximation [@maruyama55], which states that over a small time interval $\dt$, the (intractable) transition density of the SDE can be approximated by
$$
\Y_{t+\dt} \mid \Y_t \approx \N\Big(\Y_t + \dr_\tth(\Y_t)\dt, \df_\tth(\Y_t)\dt\Big),
$$
with convergence to the true SDE dynamics as $\dt \to 0$.

In order to simulate data from the Lotka-Volterra SDE model, we use the function `sde.sim`.  Here we'll generate $N = 50$ observations of the process with initial values $\Y_0 = (71, 79)$, and parameter values $\tth = (.5, .0025,.3)$, with time between observations of $\dt = 1$ year.  The `dt.sim` argument to `sde.sim` specifies the internal observation time used by the Euler-Maruyama approximation.
```{r, fig.width = 10, fig.height = 5, out.width = "90%"}

# simulation parameters
theta0 <- c(alpha = .5, beta = .0025, gamma = .3) # true parameter values
x0 <- c(H = 71, L = 79) # initial SDE values
N <- 50 # number of observations
dT <- 1 # time between observations (years)

# simulate data
lvsim <- sde.sim(model = lvmod, x0 = x0, theta = theta0,
                 nobs = N-1, # N-1 steps forward
                 dt = dT,
                 dt.sim = dT/100) # internal observation time

# plot data
Xobs <- rbind(c(x0), lvsim$data) # include first observation
tseq <- (1:N-1)*dT # observation times
clrs <- c("black", "red")
par(mar = c(4, 4, 1, 0)+.1)
plot(x = 0, type = "n", xlim = range(tseq), ylim = range(Xobs),
     xlab = "Time (years)", ylab = "Population")
lines(tseq, Xobs[,"H"], type = "o", pch = 16, col = clrs[1])
lines(tseq, Xobs[,"L"], type = "o", pch = 16, col = clrs[2])
legend("topleft", legend = c("Hare", "Lynx"), fill = clrs)
```

## Inference for multivariate SDE models

Parameter inference is conducted used a well-known data augmentation scheme due to @pedersen95.  Assume that, as above, the SDE observations are evenly spaced with interobservation time $\dt$.  For any integer $m > 0$, let $\Y_{(m)} = (\rv [m,0] \Y {m,Nm})$ denote the value of the SDE at equally spaced intervals of $\dt_m = \dt/m$.  Thus, $\Y_{(1)} = \Y$ corresponds to the observed data, and for $m > 1$, we have $\Y_{m,nm} = \Y_n$.  Thus we shall refer to the "missing data" as $\Ym = \Y_{(m)} \setminus \Y$.  The Euler-Maruyama approximation to the complete likelihood is
$$
\mathcal L(\tth \mid \Y_{(m)}) = \prod_{n=0}^{Nm-1} \varphi\Big(\Y_{m,n+1} \mid \Y_{m,n} + \dr_\tth(\Y_{m,n})\dt_m, \df_\tth(\Y_{m,n}\dt)\Big),
$$
where $\varphi(\y \mid \mathbf \mu, \mathbf \Sigma)$ is the PDF of $\y \sim \N(\mathbf \mu, \mathbf \Sigma)$.  The Bayesian data augmentation scheme then consists of chosing a prior $\pi(\tth)$ and sampling from the posterior distribution
$$
p_m(\tth \mid \Y) = \int \mathcal L(\tth \mid \Y_{(m)}) \times \pi(\tth) \, \ud \Ym.
$$
As $m \to \infty$ this approximate posterior converges to the true SDE posterior $p(\tth \mid \Y)$.

Posterior sampling from the Euler-Maruyama posterior is accomplished with the function `sde.post`.  In the following example we use $m = 1$, i.e., there is no missing data.  We'll use a Lebesgue prior $\pi(\tth) \propto 1$; more information on the [default](#defprior) and [custom](#custprior) prior specifications can be found in the following sections.
```{r, fig.width = 10, fig.height = 4, out.width = "90%"}
# initialize the posterior sampler
init <- sde.init(model = lvmod, x = Xobs, dt = dT,
                 m = 1, theta = c(.1, .1, .1))

nsamples <- 2e4
burn <- 2e3
lvpost <- sde.post(model = lvmod, init = init,
                   hyper = NULL, #prior specification
                   nsamples = nsamples, burn = burn)

# posterior histograms
tnames <- expression(alpha, beta, gamma)
par(mfrow = c(1,3))
for(ii in 1:lvmod$nparams) {
  hist(lvpost$params[,ii], breaks = 25, freq = FALSE,
       xlab = tnames[ii],
       main = parse(text = paste0("p[1](", tnames[ii], "*\" | \"*bold(Y))")))
  # superimpose true parameter value
  abline(v = theta0[ii], lwd = 4, lty = 2)
}
```

### Missing data specification with `sde.init`

In the example above there was no missing data, i.e., $m = 1$ and both components of the SDE are observed at each time $\rv [0] t n$.  In order to refine the Euler-Maruyama approximation, we simply pass a larger value of $m$ to `sde.init`:
```{r, eval = FALSE}
# 3 missing data points between each observation, so dt_m = dt/4
m <- 4
init <- sde.init(model = lvmod, x = Xobs, dt = dT,
                 m = m, theta = c(.1, .1, .1))
```
We can also assume that only the first $q < d$ components of the SDE are observed, with the last $d-q$ being latent.  In this case with $q = 1$, the lynx population would be unobserved, and this is specified with the `nvar.obs` argument:
```{r, eval = FALSE}
init <- sde.init(model = lvmod, x = Xobs, dt = dT,
                 nvar.obs = 1, # number of "observed" variables per timepoint
                 m = m, theta = c(.1, .1, .1))
```
Note that the initial data `x` must still be supplied as an $(N+1) \times d$ matrix, with the missing values corresponding to initial values for the MCMC sampler.

### Default prior specification {#defprior}

Since **`msde`** allows for some of the $q$ components of $\Y_t$ to be latent, a prior must be specified not only for $\tth$ but also for the latent variables in the initial observation $\Y_0$.

In the example above, we assumed a Lebesgue prior $\pi(\tth, \Y_0) \propto 1$, with the restriction that $\tth, \Y_0 > 0$ (as specified in the `sdeModel` class definition via the `isValidData` and `isValidParams` validators).

#### Default prior

The default prior in **`msde`** is a multivariate normal, for which the (fixed) hyper-parameters are supplied via the `hyper` argument to `sde.post`.  The `hyper` argument can either be `NULL`, or a list with elements `mu` and `Sigma`.  These consist of a named vector named matrix specifying the mean and variance of the named elements.  Unnamed elements are given a Lebesgue prior.  So for example, posterior inference for the dataset above with $m = 1$, latent variable $L$, and prior distribution
$$
L_0, \alpha, \gamma \iid \N(1, 1), \qquad \pi(H_0, \beta \mid H_0, \alpha, \gamma) \propto 1
$$
is obtained as follows:
```{r, fig.width = 10, fig.height = 4, out.width = "90%"}
# prior specification
pnames <- c("L", "alpha", "gamma")
hyper <- list(mu = rep(1, 3), Sigma = diag(3))
names(hyper$mu) <- pnames
dimnames(hyper$Sigma) <- list(pnames, pnames)

# initialize the posterior sampler
init <- sde.init(model = lvmod, x = Xobs, dt = dT,
                 m = 1, nvar.obs = 1, # L is latent
                 theta = c(.1, .1, .1))

nsamples <- 2e4
burn <- 2e3
lvpost <- sde.post(model = lvmod, init = init,
                   hyper = hyper, #prior specification
                   nsamples = nsamples, burn = burn)

# posterior histograms
tnames <- expression(alpha, beta, gamma)
par(mfrow = c(1,3))
for(ii in 1:lvmod$nparams) {
  hist(lvpost$params[,ii], breaks = 25, freq = FALSE,
       xlab = tnames[ii],
       main = parse(text = paste0("p[1](", tnames[ii], "*\" | \"*bold(Y))")))
  # superimpose true parameter value
  abline(v = theta0[ii], lwd = 4, lty = 2)
}
```

## Custom prior specification {#custprior}

**`msde`** provides a two-stage mechanism for specifying user-defined priors.  This is illustrated below with a simple log-normal prior on $\eeta = (\alpha, \gamma, \beta, L_0) = (\rv \eta 4)$, namely
$$
\pi(\eeta) \iff \log(\eta_i) \iid \N(\mu_i, \sigma_i^2),
$$
where the hyperparameters are $\bm \mu = (\rv \mu 4)$ and $\bm \sigma = (\rv \sigma 4)$.  Ultimately, the hyperparameters will be passed to `sde.post` as a two-element list, e.g.,
```{r, eval = FALSE}
hyper <- list(mu = c(0,0,0,0), sigma = c(1,1,1,1))
```

### The `sdePrior` class definition

The first step of the prior specification is to define it at the **C++** level, through the `sdPrior` class.  The class corresponding to the log-normal prior above is defined in the file `lotvolPrior.h` pasted below.
```{Rcpp, eval = FALSE}
#ifndef sdePrior_h
#define sdePrior_h 1

#include <Rcpp.h> // contains R's dlnorm function

// Prior for Lotka-Volterra Model

class sdePrior {
 private:
  static const int nHyper = 4; // (alpha, beta, gamma, L)
  double *mean, *sd; // log-normal mean and standard deviation vectors
 public:
  double logPrior(double *theta, double *x); // log-prior function
  sdePrior(double **phi, int nArgs, int *nEachArg); // constructor
  ~sdePrior(); // destructor
};

// constructor
inline sdePrior::sdePrior(double **phi, int nArgs, int *nEachArg) {
  // allocate memory for hyperparameters
  mean = new double[nHyper];
  sd = new double[nHyper];
  // hard-copy hyperparameters into Prior object
  for(int ii=0; ii<nHyper; ii++) {
    mean[ii] = phi[0][ii];
    sd[ii] = phi[1][ii];
  }
}

// destructor
inline sdePrior::~sdePrior() {
  // deallocate memory to avoid memory leaks
  delete [] mean;
  delete [] sd;
}

// log-prior function itself:
// independent log-normal densities for (alpha,beta,gamma,L)
inline double sdePrior::logPrior(double *theta, double *x) {
  double lpi = 0.0;
  // alpha,beta,gamma
  for(int ii=0; ii < 3; ii++) {
    lpi += R::dlnorm(theta[ii], mean[ii], sd[ii], 1);
  }
  // L
  lpi += R::dlnorm(x[1], mean[3], sd[3], 1);
  return lpi;
}

#endif
```
The meaning of each class member is as follows:

* `sdePrior`: The class constructor.  This is how the **C++** code collects the hyperparameters from **R**.  Its argument signature must be matched *exactly* and has the following meaning:
    * `phi`: A double-pointers of type `double`.  This is a simple mechanism to give the address of the elements of a list of numeric vectors, as the hyperparameters are defined on the **R** side.  So with the example above, $\sigma_2$ is pointed to by `phi[1][1]`.
	* `nArgs`: The number of elements in the hyperparameter vector, i.e., `length(hyper)`, which in this case is 2.  In this case this number is known in advance, but for greater flexibility it is determined at runtime automatically in the **C++** code from the specific value of `hyper`.
	* `nEachArg`: The length of each hyperparameter element, i.e., `sapply(hyper, length)`.  Again this is determined automatically from the **C++** code.
Note that for speed considerations, the contents of `phi` should be copied directly into the `sdePrior` object, held here in the private members `mean` and `sd`.
* `~sdePrior`: The class destructor, which is needed to deallocate the dynamic memory to prevent memory leaks.
* `logPrior`: The log-prior function itself, of which the signature must be matched *exactly*.  Its arguments correspond to $\tth$ and $\Y_0$.  In this case, we use the **C++** version of **R**'s `dlnorm` function, which is accessible by including `Rcpp.h`.

### Formatting the **R** input to the **C++** code

**C++** is much less forgiving that **R** when it comes to accepting incorrect inputs.  Thus, if a user accidently passed the hyperparameters to `sde.post` as e.g.,
```{r, eval = FALSE}
bad.hyper <- list(mean = c(0,0,0,0))
```
at best this would cause garbage MCMC output and at worst, the **R** session to terminate abruptly.  For this reason **`msde`** provides an input-checking mechanism, which can also used to format the hyperparameters into the list-of-numeric-vectors input expected by the **C++** code.  This is done by passing an appropriate input checking function to `sde.make.model` through the argument `hyper.check`.  This argument accepts a function with the *exact* signature of the example below.  In this example, `hyper` must be a list with elements `mu` and `sigma`, which are either:

1.  scalars, in which case each is replicated four times.
2.  vectors of length four, in which case the order is determined by $\eeta = (\alpha, \beta, \gamma, L_0)$.

```{r}
# must match argument signature _exactly_
lvcheck <- function(hyper, param.names, data.names) {
  if(is.null(names(hyper)) ||
     !identical(sort(names(hyper)), c("mu", "sigma"))) {
    stop("hyper must be a list with elements mu and sigma.")
  }
  mu <- hyper$mu
  if(length(mu) == 1) mu <- rep(mu, 4)
  if(!is.numeric(mu) || length(mu) != 4) {
    stop("mu must be a numeric scalar or vector of length four.")
  }
  sig <- hyper$sigma
  if(length(sig) == 1) sig <- rep(sig, 4)
  if(!is.numeric(sig) || length(sig) != 4 || !all(sig > 0)) {
    stop("sigma must be a positive scalar or vector of length four.")
  }
  list(mu, sig)
}
#lvcheck <- mvn.hyper.check
```

### Compiling and checking the prior

Now we are ready to create the `sde.model` object:
```{r}
data.names <- c("H", "L")
param.names <- c("alpha", "beta", "gamma")
lvmod2 <- sde.make.model(ModelFile = "lotvolModel.h",
                        PriorFile = "lotvolPrior.h", # prior specification
                        hyper.check = lvcheck, # prior input checking
                        data.names = data.names,
                        param.names = param.names)
```
We can also test the **C++** implementation of the prior against one written in **R**, using the **`msde`** function `sde.prior`.
```{r}
# generate some test values
nreta <- 12
x0 <- c(H = 71, L = 79)
theta0 <- c(alpha = .5, beta = .0025, gamma = .3)
X <- jit.vec(x0, nreta)
Theta <- jit.vec(theta0, nreta)
Eta <- cbind(Theta, L = X[,"L"])
nrphi <- 5
Phi <- lapply(1:nrphi, function(ii) list(mu = rnorm(4), sigma = rexp(4)))

# prior check

# R version
lpi.R <- matrix(NA, nreta, nrphi)
for(ii in 1:nrphi) {
  lpi.R[,ii] <- colSums(dlnorm(x =  t(Eta),
                               meanlog = Phi[[ii]]$mu,
                               sdlog = Phi[[ii]]$sigma, log = TRUE))
}

# C++ version
lpi.cpp <- matrix(NA, nreta, nrphi)
for(ii in 1:nrphi) {
  lpi.cpp[,ii] <- sde.prior(model = lvmod2, theta = Theta, x = X,
                            hyper = Phi[[ii]])
}

# compare
max.diff(lpi.R, lpi.cpp)

```

## Installation {#install}

The **`msde`** package requires on-the-fly **C++** compiling of user-specified models which is handled through the **R** package **`Rcpp`** [@eddelbuettel.francois11].

### Enable **C++** compiling for **R**

* For Windows, install [**`Rtools`**](https://cran.r-project.org/bin/windows/Rtools/).  Let the installer modify your system variable `Path`.
* For OS X, install the [**`Xcode`** command line tools](https://developer.apple.com/library/content/technotes/tn2339/_index.html).  To do this, open Terminal and run `xcode-select --install`.  You can simply press "Install" without obtaining the entire Xcode suite.
* For Linux, install ``build-essential`` and a recent version of **`g++`** or **`clang++`**.

To make sure the **C++** compiler is set up correctly, install the **``Rcpp``** package and from within **R** run the following:
```{r, eval = FALSE}
Rcpp::cppFunction("double AddTest(double x, double y) {return x + y;}")
AddTest(5.2, 3.4)
```
If the code compiles and outputs `r 5.2+3.4` then the **C++** compiler is interfaced with **R** correctly.

### Optimize settings for the **C++** compiler (optional)

It's possible to speed up **`msde`** by a reasonable amount by passing a few flags to the **C++** compiler.  This can be done by creating a `Makevars` file (or `Makevars.win` on Windows).  To do this, find your home folder by running the **R** command `Sys.getenv("HOME")`, and in that folder create a subfolder called `.R` containing the `Makevars` file (if it doesn't exist already).  Now add the following lines to th `.R/Makevars` file:
```{bash, eval = FALSE}
CXXFLAGS=-O3 -ffastmath
CXX=clang++
```
The first two options make the **C++** code faster and the third uses the **`clang++`** compiler instead of **`g++`**, which has better error messages (and is the default compiler on OS X).

### Enable **`OpenMP`** support (optional)

On a multicore machine, **`msde`** can parallelize some of its computations with **`OpenMP`** directives.  This can't be done through **R** on Windows and is supported by default on recent versions of **`g++/clang++`** on Linux.  For OS X, the default version of **`clang++`** does not support **`OpenMP`** but it is supported by that of the [LLVM Project](http://llvm.org/).  This can be installed through [Homebrew](https://brew.sh/).  After installing Homebrew using the instructions from the previous link, in Terminal run `brew install llvm`.  Then have **R** use LLVM's **`clang++`** compiler by setting the following in `.R/Makevars`:
```{bash, eval = FALSE}
CXXFLAGS=-I/usr/local/opt/llvm/include -O3 -ffast-math
LDFLAGS=-L/usr/local/opt/llvm/lib
CXX=/usr/local/opt/llvm/bin/clang++
```
Note that these compiler directives alone will not enable **`OpenMP`** support.  This happens at compile time by linking against `-fopenmp`, which is done internally by **`msde`**.

## References

### 2017-06-28 inst/doc/msde-quicktut.R

@@ -1,0 +1,269 @@
+## ---- eval = FALSE-------------------------------------------------------
+#  sde.drift <- function(x, theta) {
+#    dr <- c(theta[1]*x[1] - theta[2]*x[1]*x[2], # alpha * H - beta * H*L
+#        theta[2]*x[1]*x[2] - theta[3]*x[2]) # beta * H*L - gamma * L
+#    dr
+#  }
+
+## ---- eval = FALSE-------------------------------------------------------
+#  sde.diff <- function(x, theta) {
+#    df <- matrix(NA, 2, 2)
+#    df[1,1] <- theta[1]*x[1] + theta[2]*x[1]*x[2] # alpha * H + beta * H*L
+#    df[1,2] <- -theta[2]*x[1]*x[2] # -beta * H*L
+#    df[2,1] <- df[1,2] # -beta * H*L
+#    df[2,2] <- theta[2]*x[1]*x[2] + theta[3]*x[2] # beta * H*L + gamma * L
+#    df
+#  }
+
+## ------------------------------------------------------------------------
+require(msdeHeaders)
+
+# put lotvolModel.h in the working directory
+data.names <- c("H", "L")
+param.names <- c("alpha", "beta", "gamma")
+lvmod <- sde.make.model(ModelFile = "lotvolModel.h",
+                        data.names = data.names,
+                        param.names = param.names)
+
+## ------------------------------------------------------------------------
+# helper functions
+
+# random matrix of size nreps x length(x) from vector x
+jit.vec <- function(x, nreps) {
+  apply(t(replicate(n = nreps, expr = x, simplify = "matrix")), 2, jitter)
+}
+# maximum absolute and relative error between two arrays
+max.diff <- function(x1, x2) {
+  c(abs = max(abs(x1-x2)), rel = max(abs(x1-x2)/max(abs(x1), 1e-8)))
+}
+
+# R sde functions
+
+# drift and diffusion
+lv.drift <- function(x, theta) {
+  dr <- c(theta[1]*x[1] - theta[2]*x[1]*x[2], # alpha * H - beta * H*L
+          theta[2]*x[1]*x[2] - theta[3]*x[2]) # beta * H*L - gamma * L
+  dr
+}
+lv.diff <- function(x, theta) {
+  df <- matrix(NA, 2, 2)
+  df[1,1] <- theta[1]*x[1] + theta[2]*x[1]*x[2] # alpha * H + beta * H*L
[…]

### 2017-07-04 R/mvn.hyper.check.R
checked man

@@ -1,16 +1,16 @@
 #' Argument checking for the default multivariate normal prior.
 #'
-#' @param hyper The normal prior's hyperparameters: a list with elements \code{mu} and \code{Sigma}, corresponding to a named mean vector and variance matrix (see Details).
+#' @param hyper The normal prior's hyperparameters: \code{NULL}, or a list with elements \code{mu} and \code{Sigma}, corresponding to a named mean vector and variance matrix (see Details).
 #' @param param.names Vector of parameter names (see Details).
 #' @param data.names Vector of data names (see Details).
 #' @return A list with the following elements:
 #' \describe{
-#'   \item{mean}{the mean vector}
-#'   \item{cholSd}{the upper upper Cholesky factor of the variance matrix}
-#'   \item{thetaId}{the index of the corresponding variables in \code{theta}}
-#'   \item{xId}{the index of the corresponding variables in \code{x0}}
+#'   \item{\code{mean}}{The mean vector.}
+#'   \item{\code{cholSd}}{The upper upper Cholesky factor of the variance matrix.}
+#'   \item{\code{thetaId}}{The index of the corresponding variables in \code{theta}.}
+#'   \item{\code{xId}}{The index of the corresponding variables in \code{x0}.}
 #' }
-#' @details This function is not meant to be called directly by the user, but rather to parse the hyper-parameters of a "default" multivariate normal prior distribution to be passed to the C++ code in \code{\link{sde.prior}} and \code{\link{sde.post}}.  This default prior is multivariate normal on the elements of \code{(theta, x0)} specified by each of \code{names(mu)}, \code{rownames(Sigma)}, and \code{colnames(Sigma)}.  The remaining components are given Lebesgue priors.  If the names of \code{mu} and \code{Sigma} are inconsistent an error is thrown.
+#' @details This function is not meant to be called directly by the user, but rather to parse the hyper-parameters of a default multivariate normal prior distribution to be passed to the C++ code in \code{\link{sde.prior}} and \code{\link{sde.post}}.  This default prior is multivariate normal on the elements of \code{(theta, x0)} specified by each of \code{names(mu)}, \code{rownames(Sigma)}, and \code{colnames(Sigma)}.  The remaining components are given Lebesgue priors, or a full Lebesgue prior if \code{hyper == NULL}.  If the names of \code{mu} and \code{Sigma} are inconsistent an error is thrown.
 #' @export
 mvn.hyper.check <- function(hyper, param.names, data.names) {
   nparams <- length(param.names)

### 2017-07-04 R/sde.drift.R

@@ -1,10 +1,10 @@
 #' SDE drift function.
 #'
-#' Computes the SDE model's drift function given data and parameter.
+#' Computes the SDE model's drift function given data and parameter values.
 #' @param model An \code{sde.model} object.
 #' @param x A vector or matrix of data with \code{ndims} columns.
 #' @param theta A vector or matrix of parameters with \code{nparams} columns.
-#' @return A matrix with \code{ndims} columns containing the drift function evaluated at \code{x} and \code{theta}.  If either input contains invalid sde data or parameters an error is thrown.
+#' @return A matrix with \code{ndims} columns containing the drift function evaluated at \code{x} and \code{theta}.  If either input contains invalid SDE data or parameters an error is thrown.
 #' @examples
 #' \donttest{
 #' # compile model
@@ -29,29 +29,6 @@ sde.drift <- function(model, x, theta) {
   if(class(model) != "sde.model") {
     stop("model must be an sde.model object.")
   }
-  ## # model constants
-  ## ndims <- model$ndims
-  ## data.names <- model$data.names
-  ## nparams <- model$nparams
-  ## param.names <- model$param.names
-  ## # initialize
-  ## if(!is.matrix(x)) {
-  ##   x <- matrix(x, ncol = 1)
-  ## } else {
-  ##   x <- t(x)
-  ## }
-  ## if(!is.matrix(theta)) {
-  ##   theta <- matrix(theta, ncol = 1)
-  ## } else {
-  ##   theta <- t(theta)
-  ## }
-  ## if(debug) browser()
-  ## nreps <- max(ncol(x), ncol(theta))
-  ## if(ncol(x) == 1) x <- matrix(x, ndims, nreps)
-  ## if(ncol(theta) == 1) theta <- matrix(theta, nparams, nreps)
-  ## if(!all(c(ncol(x), ncol(theta)) == nreps)) {
-  ##   stop("x and theta have incompatible dimensions.")
-  ## }
   # initialize
   x <- .format.data(x, model$data.names, type = "matrix")
   theta <- .format.params(theta, model$param.names)
@@ -71,36 +48,6 @@ sde.drift <- function(model, x, theta) {
   if(!all(.is.valid.data(model, x, theta, single.x, single.theta, nreps))) {
     stop("x contains invalid sde data.")
   }
-  ## if(!all(model$is.params(thetaIn = as.double(theta),
-  ##                         nReps = as.integer(ifelse(single.theta, 1
-  ##                                                 , nreps))))) {
[…]

### 2017-07-04 R/sde.init.R

@@ -1,18 +1,18 @@
 #' MCMC initialization.
 #'
-#' Specifies the observed SDE data, interobservation times, initial parameter and missing data values for \code{\link{sde.post}}.
+#' Specifies the observed SDE data, interobservation times, initial parameter and missing data values to be supplied to \code{\link{sde.post}}.
 #' @param model An \code{sde.model} object.
 #' @param x An \code{nobs x ndims} matrix of data.
 #' @param dt A scalar or length \code{nobs-1} vector of interobservations times.
 #' @param theta A length \code{nparams} vector of parameter values.
-#' @param nvar.obs A scalar or length \code{nobs} vector of integers between 0 and \code{ndims} denoting the number of observed SDE variables in each row of \code{data}.  See Details.
+#' @param nvar.obs A scalar or length \code{nobs} vector of integers between 0 and \code{ndims} denoting the number of observed SDE variables in each row of \code{data}.  Defaults to \code{ndims}.  See Details.
 #' @param m Positive integer, such that \code{m-1} evenly-spaced missing data time points are placed between observations.  See Details.
-#' @return An \code{sde.init} object.  A list with elements:
+#' @return An \code{sde.init} object, corresponding to a list with elements:
 #' \describe{
-#'   \item{data}{An \code{ncomp x ndims} matrix of complete data, where \code{ncomp = N_m = m * (nobs-1)+1}.}
-#'   \item{dt.m}{The complete data interobservation time, \code{dt_m = dt/m}.}
-#'   \item{nvar.obs.m}{The number of variables observed per row of \code{data}.  Note that \code{nvar.obs.m[(i-1)*m+1] == nvar.obs[ii]}, and that \code{nvar.obs.m[i-1] == 0} if \code{i} is not a multiple of \code{m}.}
-#'   \item{params}{Parameter initial values.}
+#'   \item{\code{data}}{An \code{ncomp x ndims} matrix of complete data, where \code{ncomp = N_m = m * (nobs-1)+1}.}
+#'   \item{\code{dt.m}}{The complete data interobservation time, \code{dt_m = dt/m}.}
+#'   \item{\code{nvar.obs.m}}{The number of variables observed per row of \code{data}.  Note that \code{nvar.obs.m[(i-1)*m+1] == nvar.obs[ii]}, and that \code{nvar.obs.m[i-1] == 0} if \code{i} is not a multiple of \code{m}.}
+#'   \item{\code{params}}{Parameter initial values.}
 #' }
 #' @examples
 #' \donttest{

### 2017-07-04 R/sde.is.valid.R

@@ -1,9 +1,16 @@
-#' Data validator.
+#' SDE data and parameter validators.
+#'
+#' Checks whether input SDE data and parameters are valid.
 #' @param model An \code{sde.model} object.
-#' @param x A matrix of data.
-#' @param theta A length \code{nparams} vector of parameter values.
+#' @param x A length-\code{ndims} vector or \code{ndims}-column matrix of SDE data.
+#' @param theta A length-\code{nparams} vector or \code{nparams}-column of SDE parameter values.
+#' @return A logical scalar or vector indicating whether the given data/parameter pair is valid.
+#' @name sde.valid
 #' @examples
 #' \donttest{
+#' # Heston's model
+#' # valid data is: Z > 0
+#' # valid parameters are: gamma, sigma > 0, |rho| < 1, beta > .5 * sigma^2
 #' hex <- example.models("hest")
 #' hmod <- sde.make.model(ModelFile = hex$ModelFile,
 #'                        param.names = hex$param.names,
@@ -11,13 +18,21 @@
 #'
 #' theta <- c(alpha = 0.1, gamma = 1, beta = 0.8, sigma = 0.6, rho = -0.8)
 #'
-#' # example of not valid data
+#' # valid data
+#' x0 <- c(X = log(1000), Z = 0.1)
+#' sde.valid.data(model = hmod, x = x0, theta = theta)
+#'
+#' # invalid data
 #' x0 <- c(X = log(1000), Z = -0.1)
 #' sde.valid.data(model = hmod, x = x0, theta = theta)
 #'
-#' # example of valid data
-#' x0 <- c(X = log(1000), Z = 0.1)
-#' sde.valid.data(model = hmod, x = x0, theta = theta)
+#' # valid parameters
+#' theta <- c(alpha = 0.1, gamma = 1, beta = 0.8, sigma = 0.6, rho = -0.8)
+#' sde.valid.params(model = hmod, theta = theta)
+#'
+#' # invalid parameters
+#' theta <- c(alpha = 0.1, gamma = -4, beta = 0.8, sigma = 0.6, rho = -0.8)
+#' sde.valid.params(model = hmod, theta = theta)
 #' }
 #' @export
 sde.valid.data <- function(model, x, theta) {
@@ -41,24 +56,7 @@ sde.valid.data <- function(model, x, theta) {
                 nReps = as.integer(nreps))
 }
 
-#' Parameter validator.
-#' @param model An \code{sde.model} object.
[…]

### 2017-07-04 R/sde.post.R

@@ -1,94 +1,74 @@
-#' Gibbs sampler for multivariate SDE.
+#' MCMC sampler for the SDE posterior.
 #'
+#' A Metropolis-within-Gibbs sampler for the Euler-Maruyama approximation to the true posterior density.
 #' @param model An \code{sde.model} object constructed with \code{\link{sde.make.model}}.
 #' @param init An \code{sde.init} object constructed with \code{\link{sde.init}}.
 #' @param hyper The hyperparameters of the SDE prior.  See \code{\link{sde.prior}}.
-#' @param nsamples Number of iterations.
-#' @param burn Integer number of burnin samples, or fraction of \code{nsamples} to prepend as burnin.
-#' @param mwg.sd standard deviation jump size for Metropolis-within-Gibbs steps (see Details).
+#' @param nsamples Number of MCMC iterations.
+#' @param burn Integer number of burn-in samples, or fraction of \code{nsamples} to prepend as burn-in.
+#' @param mwg.sd Standard deviation jump size for Metropolis-within-Gibbs on parameters and missing components of first SDE observation (see Details).
 #' @param adapt Logical or list to specify adaptive Metropolis-within-Gibbs sampling (see Details).
 #' @param loglik.out Logical, whether to return the loglikelihood at each step.
 #' @param last.miss.out Logical, whether to return the missing sde components of the last observation.
 #' @param update.data Logical, whether to update the missing data.
-#' @param data.out A scalar, integer vector, or list of three integer vectors determines the subset of data to be returned (see Details).
+#' @param data.out A scalar, integer vector, or list of three integer vectors determining the subset of data to be returned (see Details).
 #' @param update.params Logical, whether to update the model parameters.
 #' @param fixed.params Logical vector of length \code{nparams} indicating which parameters are to be held fixed in the MCMC sampler.
 #' @param ncores If \code{model} is compiled with \code{OpenMP}, the number of cores to use for parallel processing.  Otherwise, uses \code{ncores = 1} and gives a warning.
 #' @param verbose Logical, whether to periodically output MCMC status.
[…]

### 2017-07-04 R/sde.prior.R

@@ -1,12 +1,12 @@
 #' SDE prior function.
 #'
-#' Evaluates the prior given data and parameter for SDE model; and hyperparameters for the SDE prior.
+#' Evaluates the SDE prior given data, parameter, and hyperparameter values.
 #' @param model An \code{sde.model} object.
 #' @param theta A vector or matrix of parameters with \code{nparams} columns.
 #' @param x A vector or matrix of data with \code{ndims} columns.
 #' @param hyper The hyperparameters of the SDE prior.  See Details.
 #' @return A vector of log-prior densities evaluated at the inputs.
-#' @details The prior is constructed at the \code{C++} level by defining a function (i.e., public member) \cr \code{double logPrior(double *theta, double *x)} within the \code{sdePrior} class (see Examples).  At the \code{R} level, the \code{hyper.check} argument of \code{sde.make.model} is a function with arguments \code{hyper}, \code{param.names}, \code{data.names} used to convert \code{hyper} into a list of \code{NULL} or double-vectors which get passed on to the \code{C++} code.  This function can also be used to throw \code{R}-level errors to protect the \code{C++} code from invalid inputs, as is done for the default prior in \code{\link{mvn.hyper.check}}.
+#' @details The prior is constructed at the \code{C++} level by defining a function (i.e., public member) \cr \code{double logPrior(double *theta, double *x)} within the \code{sdePrior} class.  At the \code{R} level, the \code{hyper.check} argument of \code{sde.make.model} is a function with arguments \code{hyper}, \code{param.names}, \code{data.names} used to convert \code{hyper} into a list of \code{NULL} or double-vectors which get passed on to the \code{C++} code.  This function can also be used to throw \code{R}-level errors to protect the \code{C++} code from invalid inputs, as is done for the default prior in \code{\link{mvn.hyper.check}}.  For a full example see the "Custom Prior" section in \code{vignette("msde-quicktut")}.
 #' @examples
 #' \donttest{
 #' hex <- example.models("hest")

### 2017-07-04 R/sde.sim.R

@@ -1,28 +1,29 @@
 #' Simulation of multivariate SDE trajectories.
 #'
+#' Simulates a discretized Euler-Maruyama approximation to the true SDE trajectory.
 #' @param model An \code{sde.model} object.
 #' @param x0 A vector or a matrix of size \code{nreps x ndims} of the SDE values at time 0.
 #' @param theta A vector or matrix of size \code{nreps x nparams} of SDE parameters.
 #' @param dt Scalar interobservation time.
-#' @param dt.sim Scalar, interobservation time for simulation.  That is, interally the interobservation time is \code{dt.sim} but only one out of every \code{dt/dt.sim} simulation steps is kept in the output.
+#' @param dt.sim Scalar interobservation time for simulation.  That is, interally the interobservation time is \code{dt.sim} but only one out of every \code{dt/dt.sim} simulation steps is kept in the output.
 #' @param nobs The number of SDE observations per trajectory to generate.
-#' @param burn Scalar burnin value.  Either an integer giving the number of burn-in steps, or a value between 0 and 1 giving the fraction of burn-in relative to \code{nobs}.
-#' @param nreps The number of trajectories to generate.
+#' @param burn Scalar burn-in value.  Either an integer giving the number of burn-in steps, or a value between 0 and 1 giving the fraction of burn-in relative to \code{nobs}.
+#' @param nreps The number of SDE trajectories to generate.
 #' @param max.bad.draws The maximum number of times that invalid forward steps are proposed.  See Details.
 #' @param verbose Whether or not to display information on the simulation.
 #' @details The simulation algorithm is a Markov process with \eqn{Y_0 = x_0} and
 #' \deqn{
-#' Y_{t+1} \sim N(Y_t + dr(Y_t, \theta) dt_{sim}, df(Y_t, \theta)df(Y_t, \theta)' dt_{sim}).
+#' Y_{t+1} \sim \mathcal{N}(Y_t + \mathrm{dr}(Y_t, \theta) dt_{\mathrm{sim}}, \mathrm{df}(Y_t, \theta) dt_{\mathrm{sim}}),
 #' }{
-#' Y_(t+1) ~ N(Y_t + dr(Y_t, theta) dt_(sim), df(Y_t, theta)df(Y_t, theta)' dt_(sim)).
+#' Y_(t+1) ~ N(Y_t + dr(Y_t, \theta) dt_(sim), df(Y_t, \theta) dt_(sim)),
 #'}
[…]

### 2017-07-04 inst/cppTemplates/sdeUtils.cpp

@@ -1,7 +1,7 @@
 #include <Rcpp.h>
 using namespace Rcpp;
 
-//[[Rcpp::depends("msdeHeaders")]]
+//[[Rcpp::depends("msde")]]
 #include <sdeMCMC.h>
 #include <mcmcUtils.h>
 

### 2017-07-04 inst/doc/msde-quicktut.R

@@ -16,7 +16,7 @@
 #  }
 
 ## ------------------------------------------------------------------------
-require(msdeHeaders)
+require(msde)
 
 # put lotvolModel.h in the working directory
 data.names <- c("H", "L")

### 2017-07-04 inst/include/ParamVanillaUpdate.h

@@ -1,7 +1,7 @@
 #ifndef ParamVanillaUpdate_h
 #define ParamVanillaUpdate_h 1
 
-//[[Rcpp::depends("msdeHeaders")]]
+//[[Rcpp::depends("msde")]]
 #include <sdeMCMC.h>
 
 // componentwise vanilla MH parameter updates

### 2017-07-04 inst/include/sdeLogLik.h

@@ -20,7 +20,7 @@ public members:
 * loglik(x, theta)
 */
 
-//[[Rcpp::depends("msdeHeaders")]]
+//[[Rcpp::depends("msde")]]
 
 #include <LinAlgUtils.h>
 #include <mvnUtils.h>

### 2017-07-04 inst/include/sdeMCMC.h

@@ -1,7 +1,7 @@
 #ifndef sdeMCMC_h
 #define sdeMCMC_h 1
 
-//[[Rcpp::depends("msdeHeaders")]]
+//[[Rcpp::depends("msde")]]
 
 // posterior inference for msde's
 

### 2017-07-04 tests/dontrun/moutest.R

@@ -1,4 +1,4 @@
-library(msdeHeaders)
+library(msde)
 source("mOU-Kalman-revised.R")
 require("mvtnorm")
 source("test-post_function.R")

### 2017-07-04 tests/dontrun/test-active.R

@@ -1,6 +1,6 @@
 #--- test active model variables -----------------------------------------------
 
-require(msdeHeaders)
+require(msde)
 #require(mvtnorm)
 
 param.names <- c("alpha", "gamma", "beta", "sigma", "rho")

### 2017-07-04 tests/dontrun/test-interface.R

@@ -14,7 +14,7 @@
 #    similarly, data.out should be scalar, vector, or list of length 3.
 
 
-require(msdeHeaders)
+require(msde)
 max.diff <- function(x1, x2) {
   c(abs = max(abs(x1-x2)), rel = max(abs(x1-x2)/max(abs(x1), 1e-8)))
 }

### 2017-07-04 tests/dontrun/test-post_function.R

@@ -1,139 +1,139 @@
-library(msdeHeaders)
+library(msde)
 source("mOU-Kalman-revised.R")
 source("r-code-parser.R")
 require("mvtnorm")
 
 # Only for mou case. This theta is one dimensional
 # Need data.names to be X_1,...,X_ndims
-test.post <- function(theta, Gamma, Lambda, Psi, init = 1,debug=FALSE){
+test.post <- function(theta, Gamma, Lambda, Psi, init = 1,debug=FALSE) {
 
-	ndims <- length(Lambda) # length of Lambda == number of dimension
+  ndims <- length(Lambda) # length of Lambda == number of dimension
 
-	# Initialize param.names
-	param.names <- names(theta)
+  # Initialize param.names
+  param.names <- names(theta)
 
-	# Initialize data.names
-	data.names <- c("X1")
-	for(nn in 2:ndims){
-		data.names <- c(data.names,paste0('X',nn))
-	}
+  # Initialize data.names
+  data.names <- c("X1")
+  for(nn in 2:ndims){
+    data.names <- c(data.names,paste0('X',nn))
+  }
 
-	# Initialize the model file. Message will confirm the conversion
-	quickParse(Gamma,Lambda,Psi)
-	
-	# Initialize model in C++
-	message("Initializing Model in C++")
-	message(".")
-	message(".")
-	message(".")
+  # Initialize the model file. Message will confirm the conversion
+  quickParse(Gamma,Lambda,Psi)
 
-	moumod <- sde.make.model(ModelFile = "mOUModel.h",
-													 data.names = data.names,
-													 param.names = param.names)
-	
-	message("Model initialization complete.")
-	message(".")
-	message(".")
-	message(".")
+  # Initialize model in C++
+  message("Initializing Model in C++")
+  message(".")
+  message(".")
+  message(".")
 
-	nreps <- 0
+  moumod <- sde.make.model(ModelFile = "mOUModel.h",
+                           data.names = data.names,
+                           param.names = param.names)
 
-	if(as.numeric(init) == init){
-		# Make a list of sde.init object
-		# generate data
-		message(init," repetitions.")
-		nreps <- init
+  message("Model initialization complete.")
+  message(".")
+  message(".")
+  message(".")
 
-		for(reps in 1:nreps){
-			message("Rep ",reps,": ")
-			message(".")
-			message(".")
-			X0 <- runif(ndims, min = -1, max = 1)
-			dT <- 1/250
+  nreps <- 0
 
[…]

### 2017-07-04 tests/dontrun/test-prior.R

@@ -3,7 +3,7 @@
 devtools::document()
 devtools::install()
 
-require(msdeHeaders)
+require(msde)
 require(mvtnorm)
 
 param.names <- c("alpha", "gamma", "beta", "sigma", "rho")

### 2017-07-04 tests/dontrun/test.R

@@ -1,6 +1,6 @@
 #--- basic tests of the msde header-only library ---------------------------
 
-require(msdeHeaders)
+require(msde)
 
 # does the package work?
 

### 2017-07-04 tests/testthat.R

@@ -1,5 +1,5 @@
 library(testthat)
-library(msdeHeaders)
+library(msde)
 
 Sys.unsetenv("R_TESTS")
-test_check("msdeHeaders")
+test_check("msde")

### 2017-07-04 tests/testthat/test-hest_diag.R

@@ -1,6 +1,6 @@
 #--- heston model on diag/sd scale ---------------------------------------------
 
-library(msdeHeaders)
+library(msde)
 
 context("heston model -- diag/sd scale")
 

### 2017-07-04 tests/testthat/test-hest_var.R

@@ -1,6 +1,6 @@
 #--- heston model on sd scale --------------------------------------------------
 
-library(msdeHeaders)
+library(msde)
 context("heston model -- var scale")
 
 # setup heston model

### 2017-07-22 old/inst/doc/msde-exmodels.Rmd
multiple models tested

---
title: "Example SDE models provided by **`msde`**"
author: "Martin Lysy, JunYong Tong, Nigel Delaney"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
bibliography: references.bib
csl: taylor-and-francis-harvard-x.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Example models provided by msde}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\tth}{\bm{\theta}}
\newcommand{\pphi}{\bm{\phi}}
\newcommand{\eeta}{\bm{\eta}}
\newcommand{\Y}{\bm{Y}}
\newcommand{\y}{\bm{y}}
\newcommand{\B}{\bm{B}}
\renewcommand{\a}{\alpha}
\renewcommand{\b}{\beta}
\newcommand{\g}{\gamma}
\newcommand{\rv}[3][1]{#2_{#1},\ldots,#2_{#3}}
\newcommand{\dr}{\bm{\Lambda}}
\newcommand{\df}{\bm{\Sigma}}
\newcommand{\dt}{\Delta t}
\newcommand{\ud}{\mathrm{d}}
\newcommand{\var}{\mathrm{var}}
\newcommand{\cov}{\mathrm{cov}}
\newcommand{\cor}{\mathrm{cor}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\Ym}{\Y_{\mathrm{miss}}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\ind}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\GGam}{\bm{\Gamma}}
\newcommand{\LLam}{\bm{\lambda}}
\newcommand{\PPsi}{\bm{\Psi}}

This vignette contains a complete description of the sample models found in `msde::example.models()`.

## Heston's stochastic volatility model

Let $S_t$ denote the value of a financial asset at time $t$.  Heston's stochastic volatility model [@heston93] is given by the pair of stochastic differential equations
$$
\begin{split}
\ud S_t & = \alpha S_t\ud t + V_t^{1/2}S_t\ud B_{1t} \\
\ud V_t & = -\gamma(V_t - \mu)\ud t + \sigma V_t^{1/2} \ud B_{2t},
\end{split}
$$
where $V_t$ is a latent stochastic volatility process, and $B_{1t}$ and $B_{2t}$ are Brownian motions with $\cor(B_{1t}, B_{2t}) = \rho$.  To improve the accuracy of the numerical discretization scheme used for inference, `msde::example.models` considers the transformed variables $X_t = \log(S_t)$ and $Z_t = 2 V_t^{1/2}$, for which Heston's SDE becomes
$$
\begin{split}
\ud X_t & = (\alpha - \tfrac 1 8 Z_t^2)\ud t + \tfrac 1 2 Z_t \ud B_{1t} \\
\ud Z_t & = (\beta/Z_t - \tfrac \gamma 2 Z_t)\ud t + \sigma \ud B_{2t},
\end{split}
$$
with $\cor(B_{1t}, B_{2t}) = \rho$.  Thus the diffusion function on the variance scale is
$$
\df_\tth(\Y_t) = \begin{bmatrix} \tfrac 1 4 Z_t^2 & \tfrac \sigma 2 Z_t \\ \tfrac \sigma 2 Z_t & \sigma^2 \end{bmatrix},
$$
where $\Y_t = (X_t, Z_t)$ and $\tth = (\alpha, \gamma, \beta, \sigma, \rho)$.  The data and parameter restrictions are $Z_t, \gamma, \sigma > 0$, $|\rho| < 1$, and $\beta > \tfrac 1 2 \sigma^2$, with the final restriction ensuring that $Z_t > 0$ with probability 1.  This model is contained in `msde::example.models(model = "hest")`.

## Bivariate Ornstein-Uhlenbeck process

This model for $\Y_t = (X_{1t}, X_{2t})$ is given by
$$
\ud \Y_t = (\GGam \Y_t + \LLam)\ud t + \PPsi \ud \bm{B}_t,
$$
where $\GGam$ is a $2\times 2$ matrix, $\LLam$ is a $2 \times 1$ vector, and $\PPsi$ is a $2\times 2$ upper Choleski factor.  The model parameters are thus $\tth = (\Gamma_{11}, \Gamma_{21}, \Gamma_{12}, \Gamma_{22}, \lambda_{1}, \lambda_2, \Psi_{11}, \Psi_{21}, \Psi_{22})$, and the model restrictions are $\Psi_{11}, \Psi_{22} > 0$.  This model is contained in `msde::example.models(model = "biou")`.

## Lotka-Volterra predator-prey model

Let $H_t$ and $L_t$ denote the number of Hare and Lynx at time $t$ coexisting in a given habitat.  The Lotka-Volterra SDE describing the interactions between these two animal populations is given by [@golightly-wilkinson10]:
$$
\begin{bmatrix} \mathrm{d} H_t \\ \mathrm{d} L_t \end{bmatrix} = \begin{bmatrix} \a H_t - \b H_tL_t \\ \b H_tL_t - \g L_t \end{bmatrix}\, \mathrm{d} t + \begin{bmatrix} \a H_t + \b H_tL_t & -\b H_tL_t \\ -\b H_tL_t & \b H_tL_t + \g L_t\end{bmatrix}^{1/2} \begin{bmatrix} \mathrm{d} B_{1t} \\ \mathrm{d} B_{2t} \end{bmatrix}.
$$
The data and parameters are all restricted to be positive.  This model is contained in `msde::example.models(model = "lotvol")`.

## Prokaryotic autoregulatory gene network model

Let $\Y_t = (R_t, P_t, Q_t, D_t)$ denote the number of molecules at time $t$ of four different compounds in an autoregulatory gene network: RNA ($R$); a functional protein ($P$); protein dimmers ($Q$); and DNA ($D$). Then @golightly.wilkinson05 define an SDE describing the dynamics of $\Y_t$ with drift and (variance-scale) diffusion functions
$$
\begin{split}
\dr_\tth(\Y_t) & =
\begin{bmatrix}
\gamma_3 D_t - \gamma_7 R_t \\
2 \gamma_6 Q_t - \gamma_8P_t + \gamma_4 R_t -\gamma_5 P_t(P_t-1) \\
\gamma_2(10-D_t) - \gamma_1 D_t Q_t - \gamma_6 Q_t + \tfrac 1 2 \gamma_6 P_t(P_t-1) \\
\gamma_2(10-D_t) - \gamma_1 D_t Q_t
\end{bmatrix} \\
\df_\tth(\Y_t) & =
\begin{bmatrix}
\gamma_3 D_t + \gamma_7 R_t & 0 & 0 & 0 \\
0 &  \gamma_8P_t + 4\gamma_6 Q_t + \gamma_4 R_t + 2 \gamma_5 P_t(P_t-1) & -2 \gamma_6 Q_t - \gamma_5 P_t(P_t-1) & 0 \\
0 & -2 \gamma_6 Q_t - \gamma_5 P_t(P_t-1) & A + \gamma_6 Q_t + \tfrac 1 2 \gamma_5 P_t(P_t-1) & A_t \\
0 & 0 & A_t & A_t
\end{bmatrix},
\end{split}
$$
where $A_t = \gamma_1D_tQ_t + \gamma_2(10-D_t)$ and $\tth = (\rv \theta 8)$, $\theta_i = \log(\gamma_i)$, are various reaction rates.  The data and parameter restrictions for this model are $\tth \in \mathbb R^8$, $\Y_t > 1$, and $D_t < 10$.  This model is contained in `msde::example.models(model = "pgnet")`.

## References

### 2017-07-22 R/Methods.R

@@ -1,0 +1,31 @@
+#' R Methods
+#'
+#' @name Methods
+#' @export
+Robj.eval <- function(obj, y) {
+  if(class(obj) != "Robj") stop("obj must be an Robj object.")
+  .Cobj_eval(obj$ptr, y)
+}
+
+#' @name Methods
+#' @export
+Robj.set.x <- function(obj, x) {
+  if(class(obj) != "Robj") stop("obj must be an Robj object.")
+  .Cobj_setx(obj$ptr, x)
+}
+
+#' @name Methods
+#' @export
+Robj.get.x <- function(obj) {
+  if(class(obj) != "Robj") stop("obj must be an Robj object.")
+  .Cobj_getx(obj$ptr)
+}
+
+#' @name Methods
+#' @export
+sde.size <- function(model) {
+  if(class(model) != "sde.model") stop("model must be an sde.model object.")
+  nparams <- .sde_nParams(model$ptr)
+  ndims <- .sde_nDims(model$ptr)
+  c(nparams = nparams, ndims = ndims)
+}

### 2017-07-22 R/RcppExports.R

@@ -1,0 +1,55 @@
+# Generated by using Rcpp::compileAttributes() -> do not edit by hand
+# Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
+
+.Cobj_eval <- function(Cobj_ptr, y) {
+    .Call(`_msde2_Cobj_eval`, Cobj_ptr, y)
+}
+
+.Cobj_setx <- function(Cobj_ptr, x) {
+    invisible(.Call(`_msde2_Cobj_setx`, Cobj_ptr, x))
+}
+
+.Cobj_getx <- function(Cobj_ptr) {
+    .Call(`_msde2_Cobj_getx`, Cobj_ptr)
+}
+
+.sde_nParams <- function(sdeptr) {
+    .Call(`_msde2_sde_nParams`, sdeptr)
+}
+
+.sde_nDims <- function(sdeptr) {
+    .Call(`_msde2_sde_nDims`, sdeptr)
+}
+
+.sde_isData <- function(sdeptr, xIn, thetaIn, singleX, singleTheta, nReps) {
+    .Call(`_msde2_isData`, sdeptr, xIn, thetaIn, singleX, singleTheta, nReps)
+}
+
+.sde_isParams <- function(sdeptr, thetaIn, nReps) {
+    .Call(`_msde2_isParams`, sdeptr, thetaIn, nReps)
+}
+
+.sde_Drift <- function(sdeptr, xIn, thetaIn, singleX, singleTheta, nReps) {
+    .Call(`_msde2_Drift`, sdeptr, xIn, thetaIn, singleX, singleTheta, nReps)
+}
+
+.sde_Diff <- function(sdeptr, xIn, thetaIn, singleX, singleTheta, nReps) {
+    .Call(`_msde2_Diff`, sdeptr, xIn, thetaIn, singleX, singleTheta, nReps)
+}
+
+.sde_Loglik <- function(sdeptr, xIn, dTIn, thetaIn, nComp, nReps, singleX, singleTheta, nCores) {
+    .Call(`_msde2_LogLik`, sdeptr, xIn, dTIn, thetaIn, nComp, nReps, singleX, singleTheta, nCores)
+}
+
+.sde_Prior <- function(sdeptr, thetaIn, xIn, singleTheta, singleX, nReps, phiIn) {
+    .Call(`_msde2_Prior`, sdeptr, thetaIn, xIn, singleTheta, singleX, nReps, phiIn)
+}
+
[…]

### 2017-07-22 R/msde.R

@@ -2,9 +2,7 @@
 #'
 #' @docType package
 #' @name msde
-#' @importFrom methods formalArgs
-#' @importFrom Rcpp sourceCpp
-#' @importFrom stats approx
+#' @useDynLib msde2, .registration = TRUE
 #' @details See package vignettes; \code{vignette("msde-quicktut")} for a tutorial and \code{vignette("msde-exmodels")} for several example models.
 #' @examples
 #' \donttest{

### 2017-07-22 R/old/sde.make.model.R

@@ -1,123 +1,0 @@
-#'@name sde.make.model
-#'@title Model Creation
-#'@description Calling \code{sde.make.model} will compile the c++ code provided in its parameters along with
-#'             c++ code included in this package in order to produce a shared library.
-#'             The functions of this shared library should be accessed by passing the result of \code{sde.make.model}
-#'             into any of the functions \code{sde.sim}, \code{sde.post}, \code{sde.diff}, \code{sde.drift},
-#'             or \code{sde.loglik}
-#'@param list \code{NULL} by default. Rather than passing model parameters to \code{sde.make.model} individually,
-#'             the parameters may be passed as a list. Any combination of model parameters may be provided in this
-#'             list (excluding debug and ...).
-#'             Values from this list will only be used if no coresponding value is provided as a parameter to \code{sde.make.model}
-#'@param ndims an integer giving the number of dimensions for the model
-#'@param nparams an integer giving the number of parameters for the model
-#'@param definitionFilePath The full file path of a file containing c++ definitions for \code{sdeDr}, \code{sdeDf},
-#'                          \code{isValidData} and \code{isValidParams}\cr
-#'                          If \code{definitionFilePath} is provided, any values for \code{sdeDr}, \code{sdeDf},
-#'                          \code{isValidData} or \code{isValidParams} will be ignored.
-#'                          The file must contain definitions for all 4 functions.
-#'@param sdeDr A string of c++ code giving a function definition for the model's drift function. \cr
-#'             The function's signature must be: void sdeDr(double dr[], double x[], double t, double params[])
-#'@param sdeDf A string of c++ code giving a function definition for the model's diffusion function. \cr
-#'             The function's signature must be: void sdeDf(double df[], double x[], double sqrtT, double params[])
-#'@param isValidData A string of c++ code giving a function definition for the model's data validation function.\cr
-#'             The function's signature must be: int isValidData(double x[]) \cr
-#'             The return value must be 1 for valid data, 0 for invalid data
-#'@param isValidParams A string of c++ code giving a function definition for the model's parameter validation function. \cr
[…]

### 2017-07-22 R/sde.diff.R

@@ -51,11 +51,11 @@ sde.diff <- function(model, x, theta) {
   if(!all(.is.valid.data(model, x, theta, single.x, single.theta, nreps))) {
     stop("x contains invalid sde data.")
   }
-  df <- model$diff(xIn = as.double(x),
-                   thetaIn = as.double(theta),
-                   singleX = as.logical(single.x),
-                   singleTheta = as.logical(single.theta),
-                   nReps = as.integer(nreps))
+  df <- .sde_Diff(sdeptr = model$ptr, xIn = as.double(x),
+                  thetaIn = as.double(theta),
+                  singleX = as.logical(single.x),
+                  singleTheta = as.logical(single.theta),
+                  nReps = as.integer(nreps))
   df <- matrix(df, nrow = nreps, ncol = ndims^2, byrow = TRUE)
   # put zeros into the off-triangular elements
   df[,lower.tri(diag(ndims))] <- 0

### 2017-07-22 R/sde.init.R

@@ -14,6 +14,7 @@
 #'   \item{\code{nvar.obs.m}}{The number of variables observed per row of \code{data}.  Note that \code{nvar.obs.m[(i-1)*m+1] == nvar.obs[ii]}, and that \code{nvar.obs.m[i-1] == 0} if \code{i} is not a multiple of \code{m}.}
 #'   \item{\code{params}}{Parameter initial values.}
 #' }
+#' @importFrom stats approx
 #' @examples
 #' \donttest{
 #' # compile Heston's model

### 2017-07-22 R/sde.make.model.R

@@ -10,33 +10,19 @@
 #' @param ... additional arguments to \code{Rcpp::sourceCpp} for compiling the C++ code.
 #'@return An \code{sde.model} object, consisting of a list with the following elements:
 #' \describe{
+#' \item{\code{model.ptr}}{Pointer to C++ sde object (\code{sdeCobj}) implementing the member functions: drift/diffusion, data/parameter validators, loglikelihood, prior distribution, forward simulation, MCMC algorithm for Bayesian inference.}
 #' \item{\code{ndims, nparams}}{The number of SDE components and parameters.}
 #' \item{\code{data.names, param.names}}{The names of the SDE components and parameters.}
-#' \item{\code{drift, diff}}{SDE drift and diffusion functions.}
-#' \item{\code{loglik}}{Loglikelihood function.}
-#' \item{\code{logprior}}{SDE prior function.}
-#' \item{\code{sim}}{Function for simulating SDE data.}
-#' \item{\code{post}}{MCMC sampler for the posterior distribution.}
 #' \item{\code{omp}}{A logical flag for whether or not the model was compiled for multicore functionality with \code{OpenMP}.}
 #' }
-#' @details The functions \code{sim}, \code{post}, \code{drift}, \code{diff}, \code{logpior}, and \code{loglik} should never be called directly. Instead use \code{sde.sim}, \code{sde.post} \code{sde.diff}, \code{sde.drift} and \code{sde.loglik}.
-#'
-#' The code is compiled by copying the \code{ModelFile} to the \code{tmpdir} directory, along with a wrapper \code{.cpp} file to be compiled by \code{Rcpp::sourceCpp}.
-#'
-#' One way to crash the R session is by compiling the same model with and without OpenMP, as \code{Rcpp} will overwrite the shared object of one with the other.  For direct speed comparisons between the two versions, make a copy of the \code{sdeModel.h} header file with a different name.
-#' @examples
-#' \donttest{
-#' hex <- example.models("hest")
-#' sde.make.model(ModelFile = hex$ModelFile,
-#'                param.names = hex$param.names,
-#'                data.names = hex$data.names)
-#'
-#' }
-#'@export
[…]

### 2017-07-22 R/zzz.R

@@ -1,6 +1,8 @@
 .onLoad <- function(libname, pkgname) {
-  assign(".msde_src_path", file.path(libname, pkgname, "msde", "src"),
+  assign(".msde_src_path",
+         file.path(libname, pkgname, "msde", "src"),
          envir = parent.env(environment()))
-  assign(".msde_examples_path", file.path(libname, pkgname, "msde", "examples"),
+  assign(".msde_examples_path",
+         file.path(libname, pkgname, "msde", "examples"),
          envir = parent.env(environment()))
 }

### 2017-07-22 inst/include/sdeMCMC.h

@@ -1,7 +1,6 @@
 #ifndef sdeMCMC_h
 #define sdeMCMC_h 1
 
-//[[Rcpp::depends("msde")]]
 
 // posterior inference for msde's
 
@@ -10,31 +9,46 @@
 // of the same object.
 // the object will inherit from sdeLogLik.
 
-#include <mvnUtils.h>
-#include <sdeLogLik.h>
-#include "sdePrior.h"
+//[[Rcpp::depends("msde")]]
+#include "mvnUtils.h"
+#include "sdeLogLik.h"
+//#include "mcmcUtils.h"
+//#include "sdePrior.h"
 
-class sdeMCMC : public sdeLogLik {
-  // inherited from sdeLogLik: nDims2, nCores
+template <class sMod, class sPi>
+class sdeMCMC : public sdeLogLik<sMod> {
+  //int nDims2, nCores; // inherited from sdeLogLik
+  using sdeLogLik<sMod>::nDims2;
+  using sdeLogLik<sMod>::nCores;
   int *missInd;
   int nMiss, nMiss0, nMissN;
-  sdePrior *prior;
+  sPi *prior;
   void mvEraker(double *mean, double *sd,
 		double *x0, double *x2,
 		double b, double b2, double *theta,
-		sdeModel *sde);
+		sMod *sde);
  public:
-  //int nComp, nDims, nParams; coming from sdeLogLik
+  //int nComp, nDims, nParams; // inherited from sdeLogLik
+  using sdeLogLik<sMod>::nComp;
+  using sdeLogLik<sMod>::nDims;
+  using sdeLogLik<sMod>::nParams;
   double *currFull, *propFull, *propAccept;
   double *currX, *propX, *currTheta, *propTheta;
   double *propU; // for acceptance rates (predrawn in parallel version)
-  //double *dT, *sqrtDT; coming from sdeLogLik
+  //double *dT, *sqrtDT; // inherited from sdeLogLik
+  using sdeLogLik<sMod>::dT;
+  using sdeLogLik<sMod>::sqrtDT;
   double *B, *sqrtB;
   int *nObsComp;
   bool *fixedTheta;
-  // propMV **mvX; coming from sdeLogLik
-  //propMV *mvTheta; don't need
-  //sdeModel *sde; coming from sdeLogLik
+  //sMod *sde; // inherited sdeLogLik
+  using sdeLogLik<sMod>::sde;
[…]

### 2017-07-22 inst/include/sdeRUtils.h

@@ -1,0 +1,167 @@
+#ifndef sdeRUtils_h
+#define sdeRUtils_h
+
+#include <Rcpp.h>
+using namespace Rcpp;
+#include "sdeUtils.h"
+#include "sdeLogLik.h"
+#include "sdeInterface.h"
+
+template <class sMod, class sPi>
+  inline int sdeRobj<sMod, sPi>::get_nDims() {
+  return sMod::nDims;
+}
+
+
+template <class sMod, class sPi>
+  inline int sdeRobj<sMod, sPi>::get_nParams() {
+  return sMod::nParams;
+}
+
+template <class sMod, class sPi>
+  inline LogicalVector sdeRobj<sMod, sPi>::isData(NumericVector xIn,
+						  NumericVector thetaIn,
+						  bool singleX,
+						  bool singleTheta, int nReps) {
+  int nDims = sMod::nDims;
+  int nParams = sMod::nParams;
+  double *x = REAL(xIn);
+  double *theta = REAL(thetaIn);
+  LogicalVector validOut(nReps);
+  sMod sde;
+  for(int ii = 0; ii < nReps; ii++) {
+    validOut[ii] = sde.isValidData(&x[ii*(!singleX)*nDims],
+				   &theta[ii*(!singleTheta)*nParams]);
+  }
+  return validOut;
+}
+
+template <class sMod, class sPi>
+  inline LogicalVector sdeRobj<sMod, sPi>::isParams(NumericVector thetaIn, int nReps) {
+  int nParams = sMod::nParams;
+  double *theta = REAL(thetaIn);
+  LogicalVector validOut(nReps);
+  sMod sde;
+  for(int ii = 0; ii < nReps; ii++) {
+    validOut[ii] = sde.isValidParams(&theta[ii*nParams]);
+  }
+  return validOut;
+}
+
+// Best to parallelize these from within R: easier and prob faster since
+// memory can be allocated in parallel there
+template <class sMod, class sPi>
+  inline NumericVector sdeRobj<sMod, sPi>::Drift(NumericVector xIn,
+						 NumericVector thetaIn,
+						 bool singleX,
+						 bool singleTheta,
+						 int nReps) {
+  int nDims = sMod::nDims;
+  int nParams = sMod::nParams;
+  double *x = REAL(xIn);
[…]

### 2017-07-22 inst/msde/src/mvnPrior.h

@@ -2,11 +2,9 @@
 #define sdePrior_h 1
 
 //[[Rcpp::depends("msde")]]
-
+#include <mvnUtils.h> // this is specific to mvnPrior
 #include "sdeModel.h"
 
-// this is specific to mvnPrior
-#include <mvnUtils.h>
 
 class sdePrior {
  private:

### 2017-07-22 inst/msde/src/sdeMakeModel.cpp

@@ -1,0 +1,13 @@
+#include <Rcpp.h>
+using namespace Rcpp;
+//[[Rcpp::depends("msde2")]]
+#include <sdeInterface.h>
+#include "sdeModel.h"
+#include "sdePrior.h"
+
+//[[Rcpp::export(".sde_MakeModel")]]
+SEXP construct() {
+  sdeCobj *sde = new sdeRobj<sdeModel, sdePrior>;
+  XPtr<sdeCobj> sdeptr(sde, true);
+  return sdeptr;
+}

### 2017-07-22 old/R/RcppExports.R

@@ -1,0 +1,11 @@
+# Generated by using Rcpp::compileAttributes() -> do not edit by hand
+# Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
+
+.sde_get_nDims <- function(sdeptr) {
+    .Call(`_msde_sde_get_nDims`, sdeptr)
+}
+
+.sde_get_nParams <- function(sdeptr) {
+    .Call(`_msde_sde_get_nParams`, sdeptr)
+}
+

### 2017-07-22 old/R/sde.make.model.R

@@ -1,0 +1,203 @@
+#' Create an SDE model object.
+#'
+#' Compiles the C++ code for various SDE-related algorithms and makes the routines available within R.
+#' @param ModelFile Path to the header file where the SDE model is defined.
+#' @param PriorFile Path to the header file where the SDE prior is defined.  See \code{\link{sde.prior}} for details.
+#' @param data.names Vector of names for the SDE components.  Defaults to \code{X1,...,Xd}.
+#' @param param.names Vector of names for the SDE parameters.  Defaults to \code{theta1,...,thetap}.
+#' @param hyper.check A function with arguments \code{hyper}, \code{param.names}, and \code{data.names} used for passing the model hyper parameters to the C++ code.  See \code{\link{mvn.hyper.check}} for details.
+#' @param OpenMP Logical; whether the model is compiled with \code{OpenMP} for C++ level parallelization.
+#' @param ... additional arguments to \code{Rcpp::sourceCpp} for compiling the C++ code.
+#'@return An \code{sde.model} object, consisting of a list with the following elements:
+#' \describe{
+#' \item{\code{model.ptr}}{Pointer to C++ sde object (\code{sdeCobj}) implementing the member functions: drift/diffusion, data/parameter validators, loglikelihood, prior distribution, forward simulation, MCMC algorithm for Bayesian inference.}
+#' \item{\code{ndims, nparams}}{The number of SDE components and parameters.}
+#' \item{\code{data.names, param.names}}{The names of the SDE components and parameters.}
+#' \item{\code{omp}}{A logical flag for whether or not the model was compiled for multicore functionality with \code{OpenMP}.}
+#' }
+#' @seealso \link{\code{sde.drift}}, \link{\code{sde.diff}}, \link{\code{sde.is.valid}}, \link{\code{sde.loglik}}, \link{\code{sde.prior}}, \link{\code{sde.sim}}, \link{\code{sde.post}}
+#'@export
+sde.make.model <- function(ModelFile, PriorFile = "default",
+                           data.names, param.names, hyper.check,
+                           OpenMP = FALSE, ..., debug = FALSE) {
+  # prior specification
[…]

### 2017-07-22 old/inst/include/sdeRUtils.h

@@ -1,0 +1,170 @@
+#ifndef sdeRUtils_h
+#define sdeRUtils_h
+
+#include <Rcpp.h>
+using namespace Rcpp;
+//#include "sdeLogLik.h"
+//[[Rcpp::depends("msde")]]
+#include "sdeInterface.h"
+//typedef sdeRobj<sMod, sPi> sRobj;
+
+template <class sMod, class sPi>
+  inline int sdeRobj<sMod, sPi>::get_nDims() {
+  return sMod::nDims;
+}
+
+
+template <class sMod, class sPi>
+  inline int sdeRobj<sMod, sPi>::get_nParams() {
+  return sMod::nParams;
+}
+
+/*
+template <class sMod, class sPi>
+  inline LogicalVector sdeRobj<sMod, sPi>::isData(NumericVector xIn,
+						  NumericVector thetaIn,
+						  bool singleX,
+						  bool singleTheta, int nReps) {
+  int nDims = sMod::nDims;
+  int nParams = sMod::nParams;
+  double *x = REAL(xIn);
+  double *theta = REAL(thetaIn);
+  LogicalVector validOut(nReps);
+  sMod sde;
+  for(int ii = 0; ii < nReps; ii++) {
+    validOut[ii] = sde.isValidData(&x[ii*(!singleX)*nDims],
+				   &theta[ii*(!singleTheta)*nParams]);
+  }
+  return validOut;
+}
+
+template <class sMod, class sPi>
+  inline LogicalVector sdeRobj<sMod, sPi>::isParams(NumericVector thetaIn, int nReps) {
+  int nParams = sMod::nParams;
+  double *theta = REAL(thetaIn);
+  LogicalVector validOut(nReps);
+  sMod sde;
+  for(int ii = 0; ii < nReps; ii++) {
+    validOut[ii] = sde.isValidParams(&theta[ii*nParams]);
+  }
+  return validOut;
+}
+
+// Best to parallelize these from within R: easier and prob faster since
+// memory can be allocated in parallel there
+template <class sMod, class sPi>
+  inline NumericVector sdeRobj<sMod, sPi>::Drift(NumericVector xIn,
+						 NumericVector thetaIn,
+						 bool singleX,
+						 bool singleTheta,
+						 int nReps) {
+  int nDims = sMod::nDims;
[…]

### 2017-07-22 old/inst/include/sdeSim.h

@@ -1,0 +1,78 @@
+#ifndef sdeSim_h
+#define sdeSim_h
+
+#include <Rcpp.h>
+using namespace Rcpp;
+#include "sdeUtils.h"
+#include "sdeInterface.h"
+
+template <class sMod, class sPi>
+  inline List sdeRobj<sMod, sPi>::Sim(int nDataOut,
+				      int N, int burn, int reps, int r,
+				      double dT, int MAXBAD,
+				      NumericVector initData,
+				      NumericVector params,
+				      bool singleX, bool singleTheta) {
+  RNGScope scope;
+
+  int nDims = sMod::nDims;
+  int nParams = sMod::nParams;
+  double sqrtDT = sqrt(dT);
+  int bad = 0;
+  // output
+  NumericVector dataOut(nDataOut);
+  int nBadDraws;
+
+  // storage
+  sMod *sde = new sMod; // sde model
+  double *mean = new double[nDims]; // mean
+  double *sd = new double[nDims*nDims]; // cholesky factor
+  double *X = new double[nDims]; // current value
+  double *tmpX = new double[nDims]; // proposed value
+  double *Z = new double[nDims]; // random draw
+  double *theta; // pointer to parameter
+  int ii,jj,kk;
+
+  for(ii = 0; ii < reps; ii++) {
+    // initialize chains
+    for(kk = 0; kk < nDims; kk++) {
+      X[kk] = initData[ii*(!singleX)*nDims + kk];
+    }
+    theta = &params[ii*(!singleTheta)*nParams];
+    // loop through
+    for(jj = -burn*r; jj < N*r; jj++) {
+      // repeatedly draw from Euler until proposal is valid
+      mvEuler<sMod>(mean, sd, X, dT, sqrtDT, theta, sde);
+      do {
+	for(kk = 0; kk < nDims; kk++) {
+	  Z[kk] = norm_rand();
+	}
+	xmvn<sMod>(X, Z, mean, sd, nDims);
+	// validate draw
+      } while(!sde->isValidData(X, theta) && bad++ < MAXBAD);
+      if (bad == MAXBAD) {
+	goto stop;
+      }
+      // store
+      if(jj >= 0 && (jj+1) % r == 0) {
+	for(kk = 0; kk < nDims; kk++) {
+	  dataOut[ii*N*nDims + (jj/r)*nDims + kk] = X[kk];
+	}
+      }
+    }
+  }
+
[…]

### 2017-07-22 old/inst/include/sdeUtils.h

@@ -1,0 +1,92 @@
+#ifndef sdeUtils_h
+#define sdeUtils_h 1
+
+//[[Rcpp::depends("msde")]]
+
+// utilities for sde's
+
+#include <LinAlgUtils.h>
+#include <mvnUtils.h>
+//#include "sdeModel.h"
+
+// xmvn, zmvn, llmvn
+template <class sMod>
+inline void xmvn(double *x, double *z,
+		 double *mean, double *sd, int n) {
+  if(sMod::diagDiff) {
+    xmvn_diag(x, z, mean, sd, n);
+  } else {
+    xmvn_chol(x, z, mean, sd, n);
+  }
+  return;
+}
+
+template <class sMod>
+inline void zmvn(double *z, double *x,
+		      double *mean, double *sd, int n, int nMax) {
+  if(sMod::diagDiff) {
+    zmvn_diag(z, x, mean, sd, n, nMax);
+  } else {
+    zmvn_chol(z, x, mean, sd, n, nMax);
+  }
+  return;
+}
+
+template <class sMod>
+inline double lmvn(double *x, double *z,
+		   double *mean, double *sd, int n) {
+  double ld;
+  if(sMod::diagDiff) {
+    ld = lmvn_diag(x, z, mean, sd, n);
+  } else {
+    ld = lmvn_chol(x, z, mean, sd, n);
+  }
+  return ld;
+}
+
+
+// rescale diffusion depending on whether it's on the sd or variance scale
+// and diagonal or not.
+template <class sMod>
+inline void scaleDiff(double *df, double sqrtDT) {
+  int ii;
+  if(sMod::diagDiff) {
+    // diagonal diffusion
+    if(sMod::sdDiff) {
+      // sd scale
+      v_mult(df, sqrtDT, sMod::nDims);
+    } else {
+      // var scale
+      for(ii=0; ii<sMod::nDims; ii++) {
+	df[ii] = sqrt(df[ii])*sqrtDT;
+      }
+    }
+  } else {
+    // matrix diffusion
+    if(!sMod::sdDiff) {
+      // var scale
+      chol_decomp(df, df, sMod::nDims);
+    }
+    U_mult(df, sqrtDT, sMod::nDims);
+  }
+  return;
+}
+
+// Euler approximation mean and variance
+template <class sMod>
+inline void mvEuler(double *mean, double *sd,
[…]

### 2017-07-22 old/inst/msde/src/sdeMakeModel.cpp

@@ -1,0 +1,17 @@
+// R object constructor
+
+#include <Rcpp.h>
+using namespace Rcpp;
+
+//[[Rcpp::depends("msde")]]
+
+#include "sdeModel.h"
+#include "sdePrior.h"
+#include <sdeInterface.h>
+
+//[[Rcpp::export(".sdeMakeModel")]]
+SEXP sdeMakeModel() {
+  sdeCobj *sde = new sdeRobj<sdeModel, sdePrior>;
+  XPtr<sdeCobj> sdeptr(sde, true);
+  return sdeptr;
+}

### 2017-07-22 old/src/RcppExports.cpp

@@ -1,0 +1,40 @@
+// Generated by using Rcpp::compileAttributes() -> do not edit by hand
+// Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
+
+#include <Rcpp.h>
+
+using namespace Rcpp;
+
+// sde_get_nDims
+int sde_get_nDims(SEXP sdeptr);
+RcppExport SEXP _msde_sde_get_nDims(SEXP sdeptrSEXP) {
+BEGIN_RCPP
+    Rcpp::RObject rcpp_result_gen;
+    Rcpp::RNGScope rcpp_rngScope_gen;
+    Rcpp::traits::input_parameter< SEXP >::type sdeptr(sdeptrSEXP);
+    rcpp_result_gen = Rcpp::wrap(sde_get_nDims(sdeptr));
+    return rcpp_result_gen;
+END_RCPP
+}
+// sde_get_nParams
+int sde_get_nParams(SEXP sdeptr);
+RcppExport SEXP _msde_sde_get_nParams(SEXP sdeptrSEXP) {
+BEGIN_RCPP
+    Rcpp::RObject rcpp_result_gen;
+    Rcpp::RNGScope rcpp_rngScope_gen;
+    Rcpp::traits::input_parameter< SEXP >::type sdeptr(sdeptrSEXP);
+    rcpp_result_gen = Rcpp::wrap(sde_get_nParams(sdeptr));
+    return rcpp_result_gen;
+END_RCPP
+}
+
+static const R_CallMethodDef CallEntries[] = {
+    {"_msde_sde_get_nDims", (DL_FUNC) &_msde_sde_get_nDims, 1},
+    {"_msde_sde_get_nParams", (DL_FUNC) &_msde_sde_get_nParams, 1},
+    {NULL, NULL, 0}
+};
+
+RcppExport void R_init_msde(DllInfo *dll) {
+    R_registerRoutines(dll, NULL, CallEntries, NULL, NULL);
+    R_useDynamicSymbols(dll, FALSE);
+}

### 2017-07-22 old/src/old/sdeUtils.cpp

@@ -1,0 +1,159 @@
+#include <Rcpp.h>
+using namespace Rcpp;
+
+#include "sdeModel.h"
+#include "sdePrior.h"
+//[[Rcpp::depends("msde")]]
+#include <sdeMCMC.h>
+#include <mcmcUtils.h>
+
+// some utilities to be exported, namely for checking the c++ code
+//
+// the function are: sde's drift, diffusion, log-likelihood, prior.
+
+
+
+//[[Rcpp::export("ndims")]]
+int sde_getNDims() {
+  return sdeModel::nDims;
+}
+
+//[[Rcpp::export("nparams")]]
+int sde_getNParams() {
+  return sdeModel::nParams;
+}
+
+//[[Rcpp::export("sde.model$is.data")]]
+LogicalVector sde_isData(NumericVector xIn, NumericVector thetaIn,
+			 bool singleX, bool singleTheta, int nReps) {
+  int nDims = sdeModel::nDims;
+  int nParams = sdeModel::nParams;
+  double *x = REAL(xIn);
+  double *theta = REAL(thetaIn);
+  LogicalVector validOut(nReps);
+  sdeModel sde;
+  for(int ii = 0; ii < nReps; ii++) {
+    validOut[ii] = sde.isValidData(&x[ii*(!singleX)*nDims],
+				   &theta[ii*(!singleTheta)*nParams]);
+  }
+  return validOut;
+}
+
+//[[Rcpp::export("sde.model$is.params")]]
+LogicalVector sde_isParams(NumericVector thetaIn, int nReps) {
+  int nParams = sdeModel::nParams;
+  double *theta = REAL(thetaIn);
+  LogicalVector validOut(nReps);
+  sdeModel sde;
+  for(int ii = 0; ii < nReps; ii++) {
+    validOut[ii] = sde.isValidParams(&theta[ii*nParams]);
+  }
+  return validOut;
+}
+
+// Best to parallelize these from within R: easier and prob faster since
+// memory can be allocated in parallel there
+//[[Rcpp::export("sde.model$drift")]]
+NumericVector sde_Drift(NumericVector xIn, NumericVector thetaIn,
+			bool singleX, bool singleTheta, int nReps) {
+  int nDims = sdeModel::nDims;
+  int nParams = sdeModel::nParams;
[…]

### 2017-07-22 old/src/sdeExports.cpp

@@ -1,0 +1,93 @@
+// Exported functions from sdeCobj
+
+#include <Rcpp.h>
+using namespace Rcpp;
+#include "sdeInterface.h"
+//#include <sdeRutils.h>
+//#include <sdeSim.h>
+//#include <sdePost.h>
+
+//[[Rcpp::export(".sde_get_nDims")]]
+int sde_get_nDims(SEXP sdeptr) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->get_nDims();
+}
+
+//[[Rcpp::export(".sde_get_nParams")]]
+int sde_get_nParams(SEXP sdeptr) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->get_nParams();
+}
+
+/*
+//[[Rcpp::export(".sde_isData")]]
+LogicalVector isData(SEXP sdeptr, NumericVector xIn, NumericVector thetaIn,
+		     bool singleX, bool singleTheta, int nReps) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->isData(xIn, thetaIn, singleX, singleTheta, nReps);
+}
+
+//[[Rcpp::export(".sde_isParams")]]
+LogicalVector isParams(SEXP sdeptr, NumericVector thetaIn, int nReps) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->isParams(thetaIn, nReps);
+}
+
+//[[Rcpp::export(".sde_Drift")]]
+NumericVector Drift(SEXP sdeptr, NumericVector xIn, NumericVector thetaIn,
+		    bool singleX, bool singleTheta, int nReps) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->Drift(xIn, thetaIn, singleX, singleTheta, nReps);
+}
+
+//[[Rcpp::export(".sde_Diff")]]
+NumericVector Diff(SEXP sdeptr, NumericVector xIn, NumericVector thetaIn,
+		   bool singleX, bool singleTheta, int nReps) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->Diff(xIn, thetaIn, singleX, singleTheta, nReps);
+}
+
+//[[Rcpp::export(".sde_Loglik")]]
+NumericVector LogLik(SEXP sdeptr, NumericVector xIn, NumericVector dTIn,
+		     NumericVector thetaIn,
+		     int nComp, int nReps,
+		     bool singleX, bool singleTheta, int nCores) {
[…]

### 2017-07-22 setup.R

@@ -3,8 +3,9 @@
 require(Rcpp)
 #require(devtools)
 
+Rcpp::compileAttributes()
 devtools::document()
-devtools::install()
+devtools::install(args = "--clean")
 
 devtools::build()
 

### 2017-07-22 src/RcppExports.cpp

@@ -1,0 +1,236 @@
+// Generated by using Rcpp::compileAttributes() -> do not edit by hand
+// Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
+
+#include <Rcpp.h>
+
+using namespace Rcpp;
+
+// Cobj_eval
+double Cobj_eval(SEXP Cobj_ptr, double y);
+RcppExport SEXP _msde2_Cobj_eval(SEXP Cobj_ptrSEXP, SEXP ySEXP) {
+BEGIN_RCPP
+    Rcpp::RObject rcpp_result_gen;
+    Rcpp::RNGScope rcpp_rngScope_gen;
+    Rcpp::traits::input_parameter< SEXP >::type Cobj_ptr(Cobj_ptrSEXP);
+    Rcpp::traits::input_parameter< double >::type y(ySEXP);
+    rcpp_result_gen = Rcpp::wrap(Cobj_eval(Cobj_ptr, y));
+    return rcpp_result_gen;
+END_RCPP
+}
+// Cobj_setx
+void Cobj_setx(SEXP Cobj_ptr, double x);
+RcppExport SEXP _msde2_Cobj_setx(SEXP Cobj_ptrSEXP, SEXP xSEXP) {
+BEGIN_RCPP
+    Rcpp::RNGScope rcpp_rngScope_gen;
+    Rcpp::traits::input_parameter< SEXP >::type Cobj_ptr(Cobj_ptrSEXP);
+    Rcpp::traits::input_parameter< double >::type x(xSEXP);
+    Cobj_setx(Cobj_ptr, x);
+    return R_NilValue;
+END_RCPP
+}
+// Cobj_getx
+double Cobj_getx(SEXP Cobj_ptr);
+RcppExport SEXP _msde2_Cobj_getx(SEXP Cobj_ptrSEXP) {
+BEGIN_RCPP
+    Rcpp::RObject rcpp_result_gen;
+    Rcpp::RNGScope rcpp_rngScope_gen;
+    Rcpp::traits::input_parameter< SEXP >::type Cobj_ptr(Cobj_ptrSEXP);
+    rcpp_result_gen = Rcpp::wrap(Cobj_getx(Cobj_ptr));
+    return rcpp_result_gen;
+END_RCPP
+}
+// sde_nParams
+double sde_nParams(SEXP sdeptr);
+RcppExport SEXP _msde2_sde_nParams(SEXP sdeptrSEXP) {
[…]

### 2017-07-22 src/sdeExports.cpp

@@ -1,0 +1,88 @@
+// generic R wrappers to exported members of instantiated Cobj
+
+#include <Rcpp.h>
+using namespace Rcpp;
+#include "sdeInterface.h"
+
+//[[Rcpp::export(".sde_nParams")]]
+double sde_nParams(SEXP sdeptr) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->get_nParams();
+}
+
+//[[Rcpp::export(".sde_nDims")]]
+double sde_nDims(SEXP sdeptr) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->get_nDims();
+}
+
+//[[Rcpp::export(".sde_isData")]]
+LogicalVector isData(SEXP sdeptr, NumericVector xIn, NumericVector thetaIn,
+		     bool singleX, bool singleTheta, int nReps) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->isData(xIn, thetaIn, singleX, singleTheta, nReps);
+}
+
+//[[Rcpp::export(".sde_isParams")]]
+LogicalVector isParams(SEXP sdeptr, NumericVector thetaIn, int nReps) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->isParams(thetaIn, nReps);
+}
+
+//[[Rcpp::export(".sde_Drift")]]
+NumericVector Drift(SEXP sdeptr, NumericVector xIn, NumericVector thetaIn,
+		    bool singleX, bool singleTheta, int nReps) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->Drift(xIn, thetaIn, singleX, singleTheta, nReps);
+}
+
+//[[Rcpp::export(".sde_Diff")]]
+NumericVector Diff(SEXP sdeptr, NumericVector xIn, NumericVector thetaIn,
+		   bool singleX, bool singleTheta, int nReps) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->Diff(xIn, thetaIn, singleX, singleTheta, nReps);
+}
+
+//[[Rcpp::export(".sde_Loglik")]]
+NumericVector LogLik(SEXP sdeptr, NumericVector xIn, NumericVector dTIn,
+		     NumericVector thetaIn,
+		     int nComp, int nReps,
+		     bool singleX, bool singleTheta, int nCores) {
+  XPtr<sdeCobj> sde(sdeptr);
+  return sde->LogLik(xIn, dTIn, thetaIn, nComp, nReps,
[…]

### 2017-07-22 tests/dontrun/DerivedA.h

@@ -1,0 +1,23 @@
+#ifndef myCobj_h
+#define myCobj_h
+
+class myCobj {
+  double x;
+public:
+  myCobj() {
+    x = 3.5;
+  }
+  double eval(double y) {
+    return x + y;
+  }
+  double get_x(void) {return x;}
+  void set_x(double _x) {
+    x = _x;
+    return;
+  }
+  ~myCobj() {
+    Rprintf("objA destroyed.\n");
+  }
+};
+
+#endif

### 2017-07-22 tests/testthat/test-hest_diag.R

@@ -1,6 +1,6 @@
 #--- heston model on diag/sd scale ---------------------------------------------
 
-library(msde)
+#library(msde)
 
 context("heston model -- diag/sd scale")
 

### 2017-07-22 tests/testthat/test-hest_sd.R

@@ -1,6 +1,6 @@
 #--- heston model on sd scale --------------------------------------------------
 
-library(msde)
+#library(msde)
 context("heston model -- sd scale")
 
 # setup heston model

### 2017-07-22 tests/testthat/test-hest_var.R

@@ -1,6 +1,6 @@
 #--- heston model on sd scale --------------------------------------------------
 
-library(msde)
+#library(msde)
 context("heston model -- var scale")
 
 # setup heston model

### 2017-07-23 R/RcppExports.R
tested Xptr/precompiled interface

@@ -1,55 +1,59 @@
 # Generated by using Rcpp::compileAttributes() -> do not edit by hand
 # Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
 
-.Cobj_eval <- function(Cobj_ptr, y) {
-    .Call(`_msde2_Cobj_eval`, Cobj_ptr, y)
+.hest_MakeModel <- function() {
+    .Call(`_msde_hestMakeModel`)
 }
 
-.Cobj_setx <- function(Cobj_ptr, x) {
-    invisible(.Call(`_msde2_Cobj_setx`, Cobj_ptr, x))
+.biou_MakeModel <- function() {
+    .Call(`_msde_biouMakeModel`)
 }
 
-.Cobj_getx <- function(Cobj_ptr) {
-    .Call(`_msde2_Cobj_getx`, Cobj_ptr)
+.pgnet_MakeModel <- function() {
+    .Call(`_msde_pgnetMakeModel`)
+}
+
+.lotvol_MakeModel <- function() {
+    .Call(`_msde_lotvolMakeModel`)
 }
 
 .sde_nParams <- function(sdeptr) {
-    .Call(`_msde2_sde_nParams`, sdeptr)
+    .Call(`_msde_sde_nParams`, sdeptr)
 }
 
 .sde_nDims <- function(sdeptr) {
-    .Call(`_msde2_sde_nDims`, sdeptr)
+    .Call(`_msde_sde_nDims`, sdeptr)
 }
 
 .sde_isData <- function(sdeptr, xIn, thetaIn, singleX, singleTheta, nReps) {
-    .Call(`_msde2_isData`, sdeptr, xIn, thetaIn, singleX, singleTheta, nReps)
+    .Call(`_msde_isData`, sdeptr, xIn, thetaIn, singleX, singleTheta, nReps)
 }
 
 .sde_isParams <- function(sdeptr, thetaIn, nReps) {
-    .Call(`_msde2_isParams`, sdeptr, thetaIn, nReps)
+    .Call(`_msde_isParams`, sdeptr, thetaIn, nReps)
 }
 
 .sde_Drift <- function(sdeptr, xIn, thetaIn, singleX, singleTheta, nReps) {
-    .Call(`_msde2_Drift`, sdeptr, xIn, thetaIn, singleX, singleTheta, nReps)
+    .Call(`_msde_Drift`, sdeptr, xIn, thetaIn, singleX, singleTheta, nReps)
 }
 
[…]

### 2017-07-23 R/sde.is.valid.R

@@ -49,7 +49,7 @@ sde.valid.data <- function(model, x, theta) {
     stop("x and theta have incompatible dimensions.")
   }
   nreps <- max(nreps)
-  .sde_isData(sdexptr = model$ptr, xIn = as.double(x),
+  .sde_isData(sdeptr = model$ptr, xIn = as.double(x),
               thetaIn = as.double(theta),
               singleX = as.logical(single.x),
               singleTheta = as.logical(single.theta),

### 2017-07-23 R/zzz.R

@@ -1,8 +1,8 @@
 .onLoad <- function(libname, pkgname) {
   assign(".msde_src_path",
-         file.path(libname, pkgname, "msde", "src"),
+         file.path(libname, pkgname, "tools"),
          envir = parent.env(environment()))
   assign(".msde_examples_path",
-         file.path(libname, pkgname, "msde", "examples"),
+         file.path(libname, pkgname, "examples"),
          envir = parent.env(environment()))
 }

### 2017-07-23 inst/include/sdeLogLik.h

@@ -20,18 +20,15 @@ public members:
 * loglik(x, theta)
 */
 
-//[[Rcpp::depends("msde")]]
-
-#include <LinAlgUtils.h>
-#include <mvnUtils.h>
-//#include "sdeModel.h"
-#include <sdeUtils.h>
+#include "LinAlgUtils.h"
+#include "mvnUtils.h"
+#include "sdeUtils.h"
 
 // parallel implementation
 #ifdef _OPENMP
 #include <omp.h>
 #else
-int omp_get_thread_num(void) {return 0;}
+inline int omp_get_thread_num(void) {return 0;}
 #endif
 
 template <class sMod>

### 2017-07-23 inst/tools/mvnPrior.h

@@ -1,0 +1,88 @@
+#ifndef sdePrior_h
+#define sdePrior_h 1
+
+//[[Rcpp::depends("msde")]]
+#include <mvnUtils.h>
+//#include "sdeModel.h"
+
+
+class sdePrior {
+ private:
+  //static const int nDims = sdeModel::nDims;
+  //static const int nParams = sdeModel::nParams;
+  int nRV, nParamRV, nDataRV; // number of active variables of each type
+  int *paramId, *dataId; // index vectors
+  double *mean, *cholSd;
+  double *tmpX, *tmpZ;
+ public:
+  double logPrior(double *theta, double *x);
+  sdePrior(double **phi, int nArgs, int *nEachArg);
+  ~sdePrior();
+};
+
+// constructor
+inline sdePrior::sdePrior(double **phi, int nArgs, int *nEachArg) {
+  int ii;
+  nRV = nEachArg[0];
+  nParamRV = nEachArg[2];
+  nDataRV = nEachArg[3];
+  if(nRV > 0) {
+    mean = new double[nRV];
+    cholSd = new double[nRV*nRV];
+    tmpX = new double[nRV];
+    tmpZ = new double[nRV];
+    // allocate mean and variance vectors
+    for(ii=0; ii<nRV; ii++) {
+      mean[ii] = phi[0][ii];
+    }
+    for(ii=0; ii<nRV*nRV; ii++) {
+      cholSd[ii] = phi[1][ii];
+    }
+    // index arrays for each type of variable
+    if(nParamRV > 0) {
+      paramId = new int[nParamRV];
+      for(ii=0; ii<nParamRV; ii++) {
+	paramId[ii] = (int) phi[2][ii] - 1;
+      }
+    }
+    if(nDataRV > 0) {
+      dataId = new int[nDataRV];
+      for(ii=0; ii<nDataRV; ii++) {
+	dataId[ii] = (int) phi[3][ii] - 1;
+      }
+    }
+  }
+}
+
+// destructor
+inline sdePrior::~sdePrior() {
+  if(nRV > 0) {
+    delete [] mean;
+    delete [] cholSd;
+    delete [] tmpX;
+    delete [] tmpZ;
+    if(nParamRV > 0) {
+      delete [] paramId;
+    }
+    if(nDataRV > 0) {
+      delete [] dataId;
+    }
+  }
+}
+
[…]

### 2017-07-23 old/vignettes/msde-quicktut copy.R

@@ -1,0 +1,27 @@
+## ---- eval = FALSE-------------------------------------------------------
+#  sde.drift <- function(x, theta) {
+#    dr <- c(theta[1]*x[1] - theta[2]*x[1]*x[2], # alpha * H - beta * H*L
+#        theta[2]*x[1]*x[2] - theta[3]*x[2]) # beta * H*L - gamma * L
+#    dr
+#  }
+
+## ---- eval = FALSE-------------------------------------------------------
+#  sde.diff <- function(x, theta) {
+#    df <- matrix(NA, 2, 2)
+#    df[1,1] <- theta[1]*x[1] + theta[2]*x[1]*x[2] # alpha * H + beta * H*L
+#    df[1,2] <- -theta[2]*x[1]*x[2] # -beta * H*L
+#    df[2,1] <- df[1,2] # -beta * H*L
+#    df[2,2] <- theta[2]*x[1]*x[2] + theta[3]*x[2] # beta * H*L + gamma * L
+#    df
+#  }
+
+## ------------------------------------------------------------------------
+require(msde)
+
+# put lotvolModel.h in the working directory
+data.names <- c("H", "L")
+param.names <- c("alpha", "beta", "gamma")
+lvmod <- sde.make.model(ModelFile = "lotvolModel.h",
+                        data.names = data.names,
+                        param.names = param.names)
+

### 2017-07-23 setup.R

@@ -5,7 +5,7 @@ require(Rcpp)
 
 Rcpp::compileAttributes()
 devtools::document()
-devtools::install(args = "--clean")
+devtools::install(args = "--install-tests")
 
 devtools::build()
 

### 2017-07-23 src/RcppExports.cpp

@@ -5,43 +5,49 @@
 
 using namespace Rcpp;
 
-// Cobj_eval
-double Cobj_eval(SEXP Cobj_ptr, double y);
-RcppExport SEXP _msde2_Cobj_eval(SEXP Cobj_ptrSEXP, SEXP ySEXP) {
+// hestMakeModel
+SEXP hestMakeModel();
+RcppExport SEXP _msde_hestMakeModel() {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
-    Rcpp::traits::input_parameter< SEXP >::type Cobj_ptr(Cobj_ptrSEXP);
-    Rcpp::traits::input_parameter< double >::type y(ySEXP);
-    rcpp_result_gen = Rcpp::wrap(Cobj_eval(Cobj_ptr, y));
+    rcpp_result_gen = Rcpp::wrap(hestMakeModel());
     return rcpp_result_gen;
 END_RCPP
 }
-// Cobj_setx
-void Cobj_setx(SEXP Cobj_ptr, double x);
-RcppExport SEXP _msde2_Cobj_setx(SEXP Cobj_ptrSEXP, SEXP xSEXP) {
-BEGIN_RCPP
-    Rcpp::RNGScope rcpp_rngScope_gen;
-    Rcpp::traits::input_parameter< SEXP >::type Cobj_ptr(Cobj_ptrSEXP);
-    Rcpp::traits::input_parameter< double >::type x(xSEXP);
-    Cobj_setx(Cobj_ptr, x);
-    return R_NilValue;
-END_RCPP
-}
-// Cobj_getx
-double Cobj_getx(SEXP Cobj_ptr);
-RcppExport SEXP _msde2_Cobj_getx(SEXP Cobj_ptrSEXP) {
+// biouMakeModel
+SEXP biouMakeModel();
+RcppExport SEXP _msde_biouMakeModel() {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
-    Rcpp::traits::input_parameter< SEXP >::type Cobj_ptr(Cobj_ptrSEXP);
-    rcpp_result_gen = Rcpp::wrap(Cobj_getx(Cobj_ptr));
+    rcpp_result_gen = Rcpp::wrap(biouMakeModel());
[…]

### 2017-07-23 src/biouModel_pkg.h

@@ -1,7 +1,7 @@
 /////////////////////////////////////////
 
-#ifndef sdeModel_h
-#define sdeModel_h 1
+#ifndef biouModel_h
+#define biouModel_h 1
 
 /////////////////////////////////////////
 
@@ -16,7 +16,8 @@
 // the parameter vector is theta = (Gamma, Lambda, lowertri(Psi)).
 
 // sde model object
-class sdeModel {
+class biouModel {
+ public:
   static const int nParams = 9; // Gamma = 4, Lambda = 2, Psi = 3
   static const int nDims = 2;
   static const bool sdDiff = true;
@@ -28,14 +29,14 @@ class sdeModel {
 };
 
 // drift function
-inline void sdeModel::sdeDr(double *dr, double *x, double *theta) {
+inline void biouModel::sdeDr(double *dr, double *x, double *theta) {
   dr[0] = theta[0]*x[0] + theta[2]*x[1] + theta[4];
   dr[1] = theta[1]*x[0] + theta[3]*x[1] + theta[5];
   return;
 }
 
 // diffusion function
-inline void sdeModel::sdeDf(double *df, double *x, double *theta) {
+inline void biouModel::sdeDf(double *df, double *x, double *theta) {
   df[0] = theta[6];
   df[1] = 0.0;
   df[2] = theta[7];
@@ -44,12 +45,12 @@ inline void sdeModel::sdeDf(double *df, double *x, double *theta) {
 }
 
 // data validator
-inline bool sdeModel::isValidData(double *x, double *theta) {
+inline bool biouModel::isValidData(double *x, double *theta) {
   return true;
 }
 
 // parameter validator
-inline bool sdeModel::isValidParams(double *theta) {
+inline bool biouModel::isValidParams(double *theta) {
   return (theta[6] > 0.0) && (theta[8] > 0.0);
 }
 

### 2017-07-23 src/hestModel_pkg.h

@@ -1,0 +1,56 @@
+/////////////////////////////////////////
+
+#ifndef hestModel_h
+#define hestModel_h 1
+
+/////////////////////////////////////////
+
+// Heston's model:
+// dXt = (alpha - .125 * Zt^2)dt + .5 * Zt dB_Xt
+// dZt = (beta/Zt - .5*gamma * Zt)dt + sigma * dB_Zt
+// cor(B_Xt, B_Zt) = rho
+
+// sde model object
+class hestModel {
+ public:
+  static const int nParams = 5;
+  static const int nDims = 2;
+  static const bool sdDiff = true;
+  static const bool diagDiff = false;
+  void sdeDr(double *dr, double *x, double *theta);
+  void sdeDf(double *df, double *x, double *theta);
+  bool isValidData(double *x, double *theta);
+  bool isValidParams(double *theta);
+};
+
+// drift function
+inline void hestModel::sdeDr(double *dr, double *x, double *theta) {
+  dr[0] = (theta[0] - .125 * x[1]*x[1]); // x
+  dr[1] = (theta[2]/x[1] - .5 * theta[1]*x[1]); // z
+  return;
+}
+
+// diffusion function
+inline void hestModel::sdeDf(double *df, double *x, double *theta) {
+  df[0] = .5 * x[1];
+  df[2] = theta[3];
+  df[3] = sqrt(1.0-theta[4]*theta[4]) * df[2];
+  df[2] *= theta[4];
+  return;
+}
+
+// data validator
+inline bool hestModel::isValidData(double *x, double *theta) {
+  return(x[1] > 0.0);
+}
+
+// parameter validator
+inline bool hestModel::isValidParams(double *theta) {
+  bool isValid;
+  isValid = (theta[1] > 0) && (theta[3] > 0);
+  isValid = isValid && (-1.0 < theta[4]) && (1.0 > theta[4]);
+  isValid = isValid && (theta[2] > 0.5 * theta[3] * theta[3]);
+  return(isValid);
+}
+
+#endif

### 2017-07-23 src/mvnPrior_pkg.h

@@ -1,27 +1,24 @@
-#ifndef sdePrior_h
-#define sdePrior_h 1
+#ifndef mvnPrior_h
+#define mvnPrior_h 1
 
-//[[Rcpp::depends("msde")]]
-#include <mvnUtils.h> // this is specific to mvnPrior
-#include "sdeModel.h"
+#include "mvnUtils.h"
 
-
-class sdePrior {
+class mvnPrior {
  private:
-  static const int nDims = sdeModel::nDims;
-  static const int nParams = sdeModel::nParams;
+  //static const int nDims = sdeModel::nDims;
+  //static const int nParams = sdeModel::nParams;
   int nRV, nParamRV, nDataRV; // number of active variables of each type
   int *paramId, *dataId; // index vectors
   double *mean, *cholSd;
   double *tmpX, *tmpZ;
  public:
   double logPrior(double *theta, double *x);
-  sdePrior(double **phi, int nArgs, int *nEachArg);
-  ~sdePrior();
+  mvnPrior(double **phi, int nArgs, int *nEachArg);
+  ~mvnPrior();
 };
 
 // constructor
-inline sdePrior::sdePrior(double **phi, int nArgs, int *nEachArg) {
+inline mvnPrior::mvnPrior(double **phi, int nArgs, int *nEachArg) {
   int ii;
   nRV = nEachArg[0];
   nParamRV = nEachArg[2];
@@ -55,7 +52,7 @@ inline sdePrior::sdePrior(double **phi, int nArgs, int *nEachArg) {
 }
 
 // destructor
-inline sdePrior::~sdePrior() {
+inline mvnPrior::~mvnPrior() {
   if(nRV > 0) {
     delete [] mean;
     delete [] cholSd;
@@ -70,7 +67,7 @@ inline sdePrior::~sdePrior() {
   }
 }
 
-inline double sdePrior::logPrior(double *theta, double *x) {
+inline double mvnPrior::logPrior(double *theta, double *x) {
   if(nRV == 0) return(0.0);
   double lp;
   int ii;

### 2017-07-23 src/pgnetModel_pkg.h

@@ -1,0 +1,75 @@
+/////////////////////////////////////////
+
+#ifndef pgnetModel_h
+#define pgnetModel_h 1
+
+/////////////////////////////////////////
+
+// Prokaryotic Autoregulatory Gene Network model
+// of Golightly-Wilkinson 2005
+
+// sde model object
+class pgnetModel {
+ public:
+  static const int nParams = 8;
+  static const int nDims = 4;
+  static const bool sdDiff = true;
+  static const bool diagDiff = false;
+  void sdeDr(double *dr, double *x, double *theta);
+  void sdeDf(double *df, double *x, double *theta);
+  bool isValidData(double *x, double *theta);
+  bool isValidParams(double *theta);
+};
+
+// drift function
+inline void pgnetModel::sdeDr(double *dr, double *x, double *theta) {
+  double K = 10.0;
+  dr[3] = exp(theta[1]) * (K - x[3]) - exp(theta[0]) * x[3] * x[2];
+  dr[1] = exp(theta[4]) * x[1] * (x[1]-1.0);
+  dr[2] = dr[3] + 0.5 * dr[1];
+  dr[0] = exp(theta[5]) * x[2];
+  dr[2] = dr[2] - dr[0];
+  dr[1] = 2 * dr[0] - dr[1] + exp(theta[3]) * x[0] - exp(theta[7]) * x[1];
+  dr[0] = exp(theta[2]) * x[3] - exp(theta[6]) * x[0];
+  return;
+}
+
+// diffusion function
+inline void pgnetModel::sdeDf(double *df, double *x, double *theta) {
+  double K = 10.0;
+  df[0] = sqrt(exp(theta[2]) * x[3] + exp(theta[6]) * x[0]);
+  df[1] = exp(theta[4]) * x[1] * (x[1]-1.0);
+  df[2] = exp(theta[5]);
+  df[5] = exp(theta[7]) * x[1] + 4*df[2] * x[2] + exp(theta[3]) * x[0] + 2.0*df[1];
+  df[14] = exp(theta[0]) * x[3] * x[2] + exp(theta[1]) * (K - x[3]);
+  df[9] = -2.0*df[2] * x[2] - df[1];
+  df[15] = df[2] * x[2] + df[14] + 0.5*df[1];
+  df[10] = df[15] - df[9] * df[9]/ df[5];
+  df[15] = sqrt(df[14] - df[14] * df[14] / df[10]);
+  df[10] = sqrt(df[10]);
[…]

### 2017-07-23 tests/testthat/test-hest_diag.R

@@ -8,6 +8,10 @@ context("heston model -- diag/sd scale")
 ModelFile <- "hestModel_diag.h"
 param.names <- c("alpha", "gamma", "beta", "sigma")
 data.names <- c("X", "Z")
+model <- sde.make.model(ModelFile = ModelFile,
+                        param.names = param.names,
+                        data.names = data.names)
+
 # heston model drift and diffusion
 drift.fun <- function(x, theta) {
   if(!is.matrix(x)) x <- t(x)

### 2017-07-23 tests/testthat/test-hest_var.R

@@ -7,6 +7,10 @@ context("heston model -- var scale")
 ModelFile <- "hestModel_var.h"
 param.names <- c("alpha", "gamma", "beta", "sigma", "rho")
 data.names <- c("X", "Z")
+model <- sde.make.model(ModelFile = ModelFile,
+                        param.names = param.names,
+                        data.names = data.names)
+
 # heston model drift and diffusion
 drift.fun <- function(x, theta) {
   if(!is.matrix(x)) x <- t(x)

### 2017-07-23 tests/testthat/test-lv_precomp.R

@@ -1,7 +1,7 @@
-#--- heston model on sd scale --------------------------------------------------
+#--- lotka-volterra model --------------------------------------------------
 
 #library(msde)
-context("heston model -- sd scale (precompiled)")
+context("lotka-volterra model -- sd scale (precompiled)")
 
 # setup heston model
 ## ModelFile <- "hestModel.h"
@@ -10,32 +10,36 @@ context("heston model -- sd scale (precompiled)")
 ## model <- sde.make.model(ModelFile = ModelFile,
 ##                         param.names = param.names,
 ##                         data.names = data.names)
-model <- sde.examples(model = "hest")
+model <- sde.examples(model = "lotvol")
 
-# heston model drift and diffusion
+# lotka-volterra model drift and diffusion
 drift.fun <- function(x, theta) {
   if(!is.matrix(x)) x <- t(x)
   if(!is.matrix(theta)) theta <- t(theta)
-  cbind(theta[,1] - .125 * x[,2]^2, theta[,3]/x[,2] - .5*theta[,2]*x[,2])
+  dr <- cbind(theta[,1]*x[,1] - theta[,2]*x[,1]*x[,2], # alpha * H - beta * H*L
+              theta[,2]*x[,1]*x[,2] - theta[,3]*x[,2]) # beta * H*L - gamma * L
+  dr
 }
 diff.fun <- function(x, theta) {
   if(!is.matrix(x)) x <- t(x)
   if(!is.matrix(theta)) theta <- t(theta)
-  cv <- .5*theta[,5]*theta[,4]*x[,2]
-  ans <- cbind(.25 * x[,2]^2, cv, cv, theta[,4]^2)
-  t(apply(ans, 1, function(x) chol(matrix(x,2,2))))
+  df <- matrix(NA, nrow(x), 4)
+  df[,1] <- theta[,1]*x[,1] + theta[,2]*x[,1]*x[,2] # alpha * H + beta * H*L
+  df[,2] <- -theta[,2]*x[,1]*x[,2] # -beta * H*L
+  df[,3] <- df[,2] # -beta * H*L
+  df[,4] <- theta[,2]*x[,1]*x[,2] + theta[,3]*x[,2] # beta * H*L + gamma * L
+  t(apply(df, 1, function(x) chol(matrix(x,2,2)))) # always use sd scale in R
 }
 # generate heston data/parameters
[…]

### 2017-07-25 inst/doc/msde-exmodels.Rmd
checked (simulatenous) OpenMP version

--- 2017-07-22 old/inst/doc/msde-exmodels.Rmd
@@
 
-This vignette contains a complete description of the sample models found in `msde::example.models()`.
+This vignette contains a complete description of the sample models found in `msde::sde.examples`.
 
@@
 where $V_t$ is a latent stochastic volatility process, and $B_{1t}$ and $B_{2t}$ are Brownian motions with $\cor(B_{1t}, B_{2t}) = \rho$.
-To improve the accuracy of the numerical discretization scheme used for inference, `msde::example.models` considers the transformed variables $X_t = \log(S_t)$ and $Z_t = 2 V_t^{1/2}$, for which Heston's SDE becomes
+To improve the accuracy of the numerical discretization scheme used for inference, `msde::sde.examples` considers the transformed variables $X_t = \log(S_t)$ and $Z_t = 2 V_t^{1/2}$, for which Heston's SDE becomes
 $$
@@
 The data and parameter restrictions are $Z_t, \gamma, \sigma > 0$, $|\rho| < 1$, and $\beta > \tfrac 1 2 \sigma^2$, with the final restriction ensuring that $Z_t > 0$ with probability 1.
-This model is contained in `msde::example.models(model = "hest")`.
+This model is contained in `msde::sde.examples(model = "hest")`.
 
@@
 The model parameters are thus $\tth = (\Gamma_{11}, \Gamma_{21}, \Gamma_{12}, \Gamma_{22}, \lambda_{1}, \lambda_2, \Psi_{11}, \Psi_{21}, \Psi_{22})$, and the model restrictions are $\Psi_{11}, \Psi_{22} > 0$.
-This model is contained in `msde::example.models(model = "biou")`.
+This model is contained in `msde::sde.examples(model = "biou")`.
 
@@
 $$
-The data and parameters are all restricted to be positive.  This model is contained in `msde::example.models(model = "lotvol")`.
+The data and parameters are all restricted to be positive.  This model is contained in `msde::sde.examples(model = "lotvol")`.
 
@@
 The data and parameter restrictions for this model are $\tth \in \mathbb R^8$, $\Y_t > 1$, and $D_t < 10$.
-This model is contained in `msde::example.models(model = "pgnet")`.
+This model is contained in `msde::sde.examples(model = "pgnet")`.
 

### 2017-07-25 R/msde.R

@@ -9,8 +9,8 @@
 #' # Posterior inference for Heston's model
 #'
 #' # compile model
-#' hfile <- sde.examples("hest", ModelFile.only = TRUE)
-#' params.names <- c("alpha", "gamma", "beta", "sigma", "rho")
+#' hfile <- sde.examples("hest", file.only = TRUE)
+#' param.names <- c("alpha", "gamma", "beta", "sigma", "rho")
 #' data.names <- c("X", "Z")
 #' hmod <- sde.make.model(ModelFile = hfile,
 #'                        param.names = param.names,

### 2017-07-25 R/sde.make.model.R

@@ -15,17 +15,17 @@
 #' \item{\code{data.names, param.names}}{The names of the SDE components and parameters.}
 #' \item{\code{omp}}{A logical flag for whether or not the model was compiled for multicore functionality with \code{OpenMP}.}
 #' }
-#' @seealso \code{\link{sde.drift}}, \code{\link{sde.diff}}, \code{\link{sde.is.valid}}, \code{\link{sde.loglik}}, \code{\link{sde.prior}}, \code{\link{sde.sim}}, \code{\link{sde.post}}.
+#' @seealso \code{\link{sde.drift}}, \code{\link{sde.diff}}, \code{\link{sde.valid}}, \code{\link{sde.loglik}}, \code{\link{sde.prior}}, \code{\link{sde.sim}}, \code{\link{sde.post}}.
 #' @importFrom Rcpp sourceCpp
 #' @importFrom methods formalArgs
 #' @importFrom tools md5sum
 #' @examples
 #' # header (C++) file for Heston's model
-#' hfile <- example.models("hest")
+#' hfile <- sde.examples("hest", file.only = TRUE)
 #' cat(readLines(hfile), sep = "\n")
 #'
 #' # compile the model
-#' params.names <- c("alpha", "gamma", "beta", "sigma", "rho")
+#' param.names <- c("alpha", "gamma", "beta", "sigma", "rho")
 #' data.names <- c("X", "Z")
 #' hmod <- sde.make.model(ModelFile = hfile,
 #'                        param.names = param.names,
@@ -35,7 +35,7 @@
 #' @export
 sde.make.model <- function(ModelFile, PriorFile = "default",
                            data.names, param.names, hyper.check,
-                           OpenMP = FALSE, ..., debug = FALSE) {
+                           OpenMP = FALSE, ...) {
   # prior specification
   if(PriorFile == "default") {
     PriorFile <- file.path(.msde_include_path, "mvnPrior.h")
@@ -54,12 +54,13 @@ sde.make.model <- function(ModelFile, PriorFile = "default",
   }
   # compile C++ code
   cppFile <- .copy.cpp.files(ModelFile, PriorFile, OpenMP)
-  if(debug) browser()
+  #if(debug) browser()
   if(OpenMP) old.env <- .omp.set()
-  sourceCpp(cppFile, env = environment(), ...)
+  fun.env <- environment()
+  sourceCpp(cppFile, env = fun.env, ...)
[…]

### 2017-07-25 tests/dontrun/test-omp.R

@@ -1,5 +1,9 @@
 #--- test omp implementation ---------------------------------------------------
 
+# check that omp = T/F models can be simultaneously loaded
+# and give same results
+
+
 devtools::document()
 devtools::install()
 
@@ -8,12 +12,21 @@ require(msde)
 # build model
 param.names <- c("alpha", "gamma", "beta", "sigma", "rho")
 data.names <- c("X", "Z")
+hfile <- sde.examples("hest", file.only = TRUE)
 
 # with omp
-hmod <- sde.make.model(ModelFile = "hestModel.h",
+hmod <- sde.make.model(ModelFile = hfile,
                        param.names = param.names,
                        data.names = data.names,
-                       OpenMP = TRUE, showOutput = TRUE, rebuild = TRUE)
+                       OpenMP = TRUE)
+
+# without omp
+hmod2 <- sde.make.model(ModelFile = hfile,
+                        param.names = param.names,
+                        data.names = data.names,
+                        OpenMP = FALSE)
+
+
 # NOTE: can't compile both versions in the same R session,
 # since Rcpp doesn't create a different .so, so get a segmentation fault.
 ## # without omp
@@ -50,7 +63,7 @@ colnames(X0) <- data.names
 # serial is better when nObs is small, presumably because memory allocation
 # is bottleneck
 
-nReps <- 1
+nReps <- 100
 nObs <- 100
 dT <- 1/252
 
@@ -65,20 +78,20 @@ hsim <- sde.sim(model = hmod, x0 = X0, theta = Theta,
                 dt = dT, dt.sim = dT, nobs = nObs, nreps = nReps)
 
 
-ncores <- 1
+# multicore
+ncores <- sample(1:8,1)
 system.time({
-  ll <- sde.loglik(model = hmod, x = hsim$data, dt = dT,
-                   theta = Theta, ncores = ncores)
+  ll1 <- sde.loglik(model = hmod, x = hsim$data, dt = dT,
+                    theta = Theta, ncores = ncores)
 })
-ll
 
+# single core
 system.time({
-  ll <- replicate(n = 100, expr = {
-    ncores <- sample(1:50, 1)
-    sde.loglik(model = hmod, x = hsim$data, dt = dT,
-               theta = Theta, ncores = ncores)
-  })
+  sde.loglik(model = hmod2, x = hsim$data, dt = dT,
+             theta = Theta)
 })
+
+range(ll1-ll2)
 
 #--- MCMC test -----------------------------------------------------------------
 
[…]

### 2018-03-03 tests/dontrun/test-pmcmc.R
edited vignettes: msde-pmcmc

@@ -95,12 +95,12 @@ mbm <- microbenchmark(
   "MCMC" = {
     bpost <- sde.post(bmod, binit, hyper = hyper,
                       fixed.params = fixed.params,
-                      nsamples = nsamples, burn = burn)
+                      nsamples = nsamples, burn = burn, verbose = FALSE)
   },
   "Particle MCMC" = {
     ppost <- sde.pmcmc(model = bmod, binit, theta0, fixed.params, nsamples, npart, dT,
                        resample = "multi", threshold = 0.5, mwg.sd)
   },
-  times = 5 # number of times to evaluate the expression
+  times = 3 # number of times to evaluate the expression
 )
-mbm
+print(mbm)

### 2018-03-03 inst/doc/msde-pmcmc.Rmd
fixed a numerical typo in benchmark study in msde-pmcmc

---
title: "Particle MCMC for Multivariate Stochastic Differential Equations with **`msde`**"
author: "Martin Lysy, Feiyu Zhu"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
bibliography: references.bib
csl: taylor-and-francis-harvard-x.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Particle MCMC with msde}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
\newcommand{\Ep}{\mathrm{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\idm}{\mathrm{I}}
\newcommand{\bm}[1]{\boldsymbol{#1}}

This vignette shows how to realize particle MCMC for multivariate SDEs with **`msde`**.

## Introduction

Parameter estimation of multivariate stochastic differential equations is a very meaningful but usually difficult task. If we translate a given stochastic differential equation into a non-linear state-space model (SSM), then Particle Markov chain Monte Carlo methods are the state-of-the-art technique that can be applied to make Bayesian inference feasible in such a situation.

For tutorial purposes, we will not delve into the theoretical details behind. Instead, we will explain the methods by directly showing the code, output and plots. A bivariate Ornstein-Uhlenbeck process will be used as an example. We will then compare the posterior given by Metropolis-within-Gibbs MCMC and that produced by particle MCMC with the analytic result given by Kalman Filter. A benchmark test will also be carried out and we will discuss the pros and cons of particle MCMC at the end.

## Multivariate Ornstein-Uhlenbeck Process
The $p$-dimensional multivariate Ornstein-Uhlenbeck (mOU) process $Y_t = (Y_{1t}, \ldots, Y_{dt})$ satisfies the SDE
$$
dY_t = (\Gamma Y_t + \Lambda)dt + \Phi^{1/2} dB_t
$$
where $B_t = (B_{1t}, \ldots, B_{pt})$ is $p$-dimensional Brownian motion.  Its Euler discretization is of the form
$$
Y_{n+1} = Y_n + (\Gamma Y_n + \Lambda) \Delta_n + \Phi^{1/2} \Delta B_n,
$$
where $Y_n = Y(t_n)$, $\Delta_n = t_{n+1} - t_n$ and 
$$
\Delta B_n = B(t_{n+1}) - B(t_n) \overset{\textit{ind}}{\sim} \mathcal N(0, \Delta_n).
$$
Thus, $Y_0, \ldots, Y_N$ is multivariate normal Markov chain for which the marginal distribution of any subset of timepoints and/or components can be efficiently calculated using the Kalman filter.  This can be used to check the MCMC output of `sde.post` and particle MCMC output as in the following example.

## MCMC for Bivariate OU Process
For reproducibility, we may first set up a seed.
```{r seed}
set.seed(123)
```

### Data simulation
We choose the pre-compiled bivariate OU model supplied by **`msde`**.
```{r bmod}
library(msde)
bmod <- sde.examples("biou")
```
Then we can generate some initial parameter values
```{r init-values}
# parameter values
Gamma0 <- .1 * crossprod(matrix(rnorm(4),2,2))
Lambda0 <- rnorm(2)
Phi0 <- crossprod(matrix(rnorm(4),2,2))
Psi0 <- chol(Phi0) # precompiled model uses the Cholesky scale
theta0 <- c(Gamma0, Lambda0, Psi0[c(1,3,4)])
names(theta0) <- bmod$param.names
# initial sde value
Y0 <- rnorm(2)
names(Y0) <- bmod$data.names
```
We set the total observation period to be `10` and simulate observations of SDE as follows:
```{r sim}
# simulation
dT <- runif(1, max = .1) # time step
nObs <- 10
bsim <- sde.sim(bmod, x0 = Y0, theta = theta0,
                dt = dT, dt.sim = dT, nobs = nObs)
YObs <- bsim$data
```
For description of the options in `sde.sim`, please check the help manual of `sde.sim`.

Finally, we can initializa the bivariate OU model.
```{r initialization}
# initialization before MCMC
binit <- sde.init(bmod, x = YObs, dt = dT, theta = theta0,
                  nvar.obs = 1) # second component is unobserved
```

### Posterior sampling via MCMC
We assume only $\Lambda_1$ is unknown (i.e., needs to be estimated). Thus, we introduce a vector `fixed.params` of logical `TRUE` and `FALSE` to indicate if a parameter is fixed or not. Since $\Lambda_1$ is unknown, it is, therefore, not fixed. 
```{r fixed-params}
# only Lambda1 is unknown
fixed.params <- rep(TRUE, bmod$nparams)
names(fixed.params) <- bmod$param.names
fixed.params["Lambda1"] <- FALSE
```
Then we set the prior on the unknown parameter(s) and observations.
```{r prior}
# prior on (Lambda1, Y_0)
hyper <- list(mu = c(0,0), Sigma = diag(2))
names(hyper$mu) <- bmod$data.names
dimnames(hyper$Sigma) <- rep(list(bmod$data.names), 2)
```
We set the total number of posterior samples as `100,000` with the burn-in period as `1,000`. The posterior sampling via MCMC is given as follows:
```{r mcmc, cache = TRUE}
# posterior sampling
nsamples <- 1e5
burn <- 1e3
bpost <- sde.post(bmod, binit, hyper = hyper,
                  fixed.params = fixed.params,
                  nsamples = nsamples, burn = burn)
```
Within `sde.post`, Metropolis-within-Gibbs (MWG) is used. The MWG jump size can be specified as a scalar, a vector or length `nparams + ndims`, or a named vector containing the elements defined by `sde.init$nvar.obs.m[1]` (the missing variables in the first SDE observation) and `fixed.params` (the SDE parameters which are not held fixed). For each MWG random variable, the default jump sizes are `.25 * |initial_value|` when `|initial_value| > 0`, and `1` otherwise. By default, `sde.post` carries out an adaptive MCMC proposal based on @adaptMCMC. At step *n* of the MCMC, the jump size of each MWG random variable is increased or decreased by $\delta(n)$, depending on whether the cumulative acceptance rate is above or below the optimal value of `0.44`.  If $\sigma_n$ is the size of the jump at step `n`, then the next jump size is determined by
$$
\log(\sigma_{n+1}) = \log(\sigma_n) \pm \delta(n), \qquad \delta(n) = \min(.01, 1/n^{1/2}).
$$
For further details of the implementation, please check the description of `sde.post`.

It is widely recognized that the optimal coordinate-wise acceptance rate should be around 44%. As we can see, our result 43.8% is very good in terms of the acceptance rate.

After the posterior sampling, we can extract the posterior samples corresponding to the unknown parameter $\Lambda_1$.
```{r L1-mcmc}
L1.mcmc <- bpost$params[,"Lambda1"]
```

### Analytic sampling via Kalman filter
In order to check the sampling result given by MCMC, we can compare it with the analytic solution given by Kalman filter.
```{r kalman, cache = TRUE}
# analytic posterior
L1.seq <- seq(min(L1.mcmc), max(L1.mcmc), len = 500)
L1.loglik <- sapply(L1.seq, function(l1) {
  lambda <- Lambda0
  lambda[1] <- l1
  mou.loglik(X = YObs, dt = dT, nvar.obs = 1,
             Gamma = Gamma0, Lambda = lambda, Phi = Phi0,
             mu0 = hyper$mu, Sigma0 = hyper$Sigma)
})

# normalize density
L1.Kalman <- exp(L1.loglik - max(L1.loglik))
L1.Kalman <- L1.Kalman/sum(L1.Kalman)/(L1.seq[2]-L1.seq[1])
```
In the following subsection, we will then compare the histogram of posterior samples given by MCMC with the analytic curve given by Kalman filter.

### Result comparison
As we can see from the following plot, the MCMC method is fine but definitely not great in this case. 
```{r mcmc-plot, fig.width = 10, fig.height = 5, out.width = "90%"}
# compare MCMC with Kalman filter
hist(L1.mcmc, breaks = 100, freq = FALSE,
     main = expression(p(Lambda[1]*" | "*bold(Y)[1])),
     xlab = expression(Lambda[1]))
lines(L1.seq, L1.Kalman, col = "red")
legend("topright", legend = c("Analytic", "MCMC"),
       pch = c(NA, 22), lty = c(1, NA), col = c("red", "black"))
```

Later we will show that even the basic particle MCMC can produce better result than the carefully tuned MWG MCMC.

## Particle MCMC for Bivariate OU Process
### Sequential Monte Carlo implementation
Particle MCMC (PMCMC) are a new class of MCMC techniques proposed in @pmcmc which rely on Sequential Monte Carlo (SMC) methods to build efficient high dimensional proposal distributions for MCMC samplers. Since SMC sampling plays a central role in the particle MCMC methodology, we abstract a group of template classes of SMC methods for SDEs in the header file `sdeSMC.h`. The header `sdeSMC.h` contains two main template classes: `sdeParticle` and `sdeFilter`. Class `sdeParticle` defines the state of particles of `sdeModel` class, comprising of the complete data at a given time point and providing storage for calculations. Class `sdeFilter` provides the filtering method based on `sdeParticle`, i.e. everything we need to calculate particle weights. In order to use these template classes in **`R`** to implement the SMC methods, we further build a header file `sdePF.h` in which the particle filtering method is a member function of `sdeRobj` class specified by `sMode` (SDE model type) and `sPi` (SDE prior type). The function header (with preprocessor directives) is given below
```{Rcpp sdePF, eval = FALSE}
#include <RcppArmadillo.h>
#include <smctc.h>
#include "sdeInterface.h"
#include "sdeSMC.h"
typedef Rcpp::LogicalVector Logical;
typedef Rcpp::NumericVector Numeric;
typedef Rcpp::IntegerVector Integer;
typedef Rcpp::NumericMatrix NumericMatrix;
typedef Rcpp::List List;

template <class sMod, class sPi>
  inline List sdeRobj<sMod, sPi>::particleEval(Numeric initParams,
                 NumericMatrix initData,
                 Numeric dT, Integer nDimsPerObs,
                 int nPart, int resample,
                 double dThreshold,
                 NumericMatrix NormalDraws,
                 bool hasNormalDraws,
                 bool historyOut)
```
For the details of the full function body, users are referred to the header file `sdePF.h`. All the SMC methods implementation in **`msde`** are specifically desinged for SDEs in mind and are based on the more general Sequential Monte Carlo Template Class (SMCTC) by @SMCTC and `RcppSMC` in **`R`** in which many low-level SMC mechanisms have been realized. Currently, `sdeSMC.h` is still under active development in aim to incorporate smoothing methods, parallel processing and the ability to deal with invalid data.

For a basic but general SMC algorithm exposition, the tutorial given by @tutorial is a good starting point. Note that particle filters for state-space models can be regarded as a special SMC algorithm. In the following subsection, we will discuss the idea and algorithm of particle MCMC for inference in state-space models.

### Particle MCMC algorithm
A general SSM with static parameters $\theta \in \Theta \subset \mathbb{R}^{n_\theta}$ (which may be multidimensional) consists of a hidden state process $\{X_n\}_{n \geq 1}$ and a observation process $\{Y_n\}_{n \geq 1}$. For inference for non-linear non-Gaussian SSMs, usual MCMC algorithms would implement some sort of Metropolis-within-Gibbs approach to sampling from the joint posterior $p(\theta, \bm{x} | \bm{y})$. However, due to the high correlation between $\theta$ and $\bm{X}$ (given $\bm{Y}$), such algorithms are very inefficient. In contrast, MCMC directly on $p(\theta | \bm{y})$ would be much more efficient, but it's impossible because the marginal isn't available in closed form for general SDEs. Particle MCMC overcomes this challenge by using an unbiased estimate, on an extended space including all the variables $(\theta, \bm{X}, \bm{Y})$, obtained by SMC. The point of particle MCMC is to be able to do MCMC on $p(\theta | \bm{y})$ directly (in terms of MCMC efficiency).

Currently, `sde.pmcmc` in **`msde`** only implements the *particle marginal Metropolis-Hastings* (PMMH) sampler in the particle MCMC toolbox which will be used for our later simulation study. PMMH is an approximation of an ideal *pseudo marginal Metropolis-Hastings* sampler (see @pseudoMCMC) for sampling from $p(\theta, x_{1:T} | y_{1:T})$. It utilizes the following proposal density 
$$
q((\theta^*, x^*_{1:T}) | (\theta, x_{1:T})) := q(\theta^* | \theta) p_{\theta^*}(x^*_{1:T} | y_{1:T})
$$
where the proposed $x^*_{1:T}$ is perfectly *adapted* to the proposed $\theta^*$. The proposal density $q(\theta^* | \theta)$ should be specified by users. In `sde.pmcmc`, a very basic IID-Metropolis proposal is implemented. Samples from $p_{\theta^*}(x^*_{1:T} | y_{1:T})$ can be obtained from an SMC algorithm targeting $p_{\theta^*}(x^*_{1:T} | y_{1:T})$. In fact we do not need the analytic expression of $p_{\theta^*}(x^*_{1:T} | y_{1:T})$ to calculate the acceptance ratio since the ratio can be transformed as
$$
\frac{p(\theta^*, x^*_{1:T} | y_{1:T})}{p(\theta, x_{1:T} | y_{1:T})} \frac{q((\theta, x_{1:T}) | (\theta^*, x^*_{1:T}))}{q((\theta^*, x^*_{1:T}) | (\theta, x_{1:T}))} = \frac{p_{\theta^*}(y_{1:T}) p(\theta^*)}{p_\theta(y_{1:T})p(\theta)} \frac{q(\theta | \theta^*)}{q(\theta^* | \theta)}
$$
In the above formula, $p_\theta(y_{1:T})$ can be replaced by its estimated expression $\hat{p}_\theta(y_{1:T})$
$$
\hat{p}_\theta(y_{1:T}) := \hat{p}_\theta(y_1) \prod_{n=2}^T \hat{p}_\theta(y_n | y_{1:n-1})
$$
which are approximated by the unnormalized particle weights (not the incremental weights) at time $T$, i.e.
$$
\hat{p}_\theta(y_{1:T}) = \frac{1}{N} \sum_{i=1}^N w(X_{1:T}^i)
$$
where $N$ is the pre-specified total number of particles.

In practice, we store the unnormalized weights at time point $n$ as $\log$-weights, i.e. $\log w_n(X_{1:n}^i)$. To avoid possible overflow problem, we had better calculate $\log \hat{p}_\theta(y_{1:n})$ by using the following formula
$$
\log \hat{p}_\theta(y_{1:n}) = \log\left( \frac{1}{N} \sum_{i=1}^N \exp\left[\log w_n(X_{1:n}^i) - C_n\right] \right) + C_n
$$
where $C_n = \max_{1\leq i \leq N}\{\log w_n(X_{1:n}^i)\}$.

### Posterior sampling via particle MCMC
Let's set the number of particles as `100` and use just IID-Metropolis proposal for updating the unknown parameter $\Lambda_1$. Here our aim is to show that even "off-the-shelf" choice can give satisfactory result.
```{r pmcmc, cache = TRUE}
npart <- 100
mwg.sd <- 0.3 # for convenience, just use a fixed jump size instead of an adaptive one
ppost <- sde.pmcmc(model = bmod, binit, theta0, fixed.params, nsamples, npart, dT,
                   resample = "multi", threshold = 0.5, mwg.sd)
```
Then we can check the acceptance rate.
```{r accept}
# check the acceptance rate
accept <- ppost$accept
print(accept)
```
The acceptance rate is 94.6% which is close to the optimal 100%.

Finally, we extract the posterior samples of $\Lambda_1$ which will be used for comparison in the following subsection.
```{r}
# posterior mean of theta
L1.pmcmc <- ppost$params[ ,!fixed.params]
```
### Result comparison 
Similarly, we compare the histogram of posterior samples of $\Lambda_1$ given by particle MCMC with the analytic curve solved by Kalman filter.
```{r pmcmc-plot, fig.width = 10, fig.height = 5, out.width = "90%"}
# compare particle MCMC with Kalman filter
hist(L1.pmcmc, breaks = 100, freq = FALSE,
     main = expression(p(Lambda[1]*" | "*bold(Y)[1])),
     xlab = expression(Lambda[1]))
lines(L1.seq, L1.Kalman, col = "red")
legend("topright", legend = c("Analytic", "PMCMC"),
       pch = c(NA, 22), lty = c(1, NA), col = c("red", "black"))
```

It is clear that the result given by particle MCMC can fit the analytic curve better than the MWG MCMC. However, such gain in performance comes at a price, as we will discuss in the next section.

## Benchmark Study
### Numerical test results
The better performance of particle MCMC sacrifices the computational speed. We can do a benchmark test to show the computational complexity of particle MCMC relative to the usual MCMC.
```{r benchmark, cache = TRUE}
require(microbenchmark)
mbm <- microbenchmark(
  "MCMC" = {
    bpost <- sde.post(bmod, binit, hyper = hyper,
                      fixed.params = fixed.params,
                      nsamples = nsamples, burn = burn, verbose = FALSE)
  },
  "PMCMC" = {
    ppost <- sde.pmcmc(model = bmod, binit, theta0, fixed.params, nsamples, npart, dT,
                       resample = "multi", threshold = 0.5, mwg.sd)
  },
  times = 50 # number of times to evaluate the expression
)
```
The benchmark results are printed as follows:
```{r, cache = TRUE}
print(mbm)
```
The time unit is "millisecond". The total number of evaluation `neval` is set to be `50`. The column `min`, `max`, `mean`, `median` represent the minimum, maximum, mean and median time spent, resprectively. Besides, `lq` and `uq` are, respectively, lower quantile and upper quantile of the sample. If the `multcomp` package is available, then a statistical ranking is calculated and displayed in compact letter display from in the `cld` column

The mean amount of time spent to run the adaptive MWG MCMC is just about 0.3 seconds. Whereas, the mean time needed to run the particle MCMC is around 50 seconds. Even though the particle MCMC shows better performance in our example above, the computational cost is much larger than the usual MCMC.

### Test results visualization
We can also visualize the benchmark results.
```{r bench-plot, cache = TRUE, fig.width = 10, fig.height = 5, out.width = "90%"}
require(ggplot2)
autoplot(mbm)
```

The plot is called violin plot which is a method of plotting numeric data. It is similar to box plot but with a rotated kernel density plot on each side. The visualization also confirms that particle MCMC is computationally more intensive than MCMC.

## References

### 2018-03-22 inst/doc/msde-pmcmc.Rmd
recompiled msde-pmcmc.Rmd

@@
 \newcommand{\bm}[1]{\boldsymbol{#1}}
+\newcommand{\N}{\mathcal{N}}
 
@@
 
-Currently, `sde.pmcmc` in **`msde`** only implements the *particle marginal Metropolis-Hastings* (PMMH) sampler in the particle MCMC toolbox which will be used for our later simulation study.
+Currently, `sde.pmcmc`^[Since `sde.pmcmc` is still under development, it is currently not available for users of **`msde`**.] in **`msde`** only implements the *particle marginal Metropolis-Hastings* (PMMH) sampler in the particle MCMC toolbox which will be used for our later simulation study.
 PMMH is an approximation of an ideal *pseudo marginal Metropolis-Hastings* sampler (see @pseudoMCMC) for sampling from $p(\theta, x_{1:T} | y_{1:T})$.
@@
 ### Posterior sampling via particle MCMC
-Let's set the number of particles as `100` and use just IID-Metropolis proposal for updating the unknown parameter $\Lambda_1$.
+Let's set the number of particles as `100` and just use an adaptive random walk proposal for updating the unknown parameter $\Lambda_1$.
 Here our aim is to show that even "off-the-shelf" choice can give satisfactory result.
+```{r source, echo = FALSE}
+source("sde.pmcmc.R")
+```
 ```{r pmcmc, cache = TRUE}
 npart <- 100
-mwg.sd <- 0.3 # for convenience, just use a fixed jump size instead of an adaptive one
-ppost <- sde.pmcmc(model = bmod, binit, theta0, fixed.params, nsamples, npart, dT,
-                   resample = "multi", threshold = 0.5, mwg.sd)
+mwg.sd <- 1
+delta <- .618
+ppost <- sde.pmcmc(model = bmod, binit, theta0, fixed.params, hyper,
+                   nsamples, npart, dT,
+                   resample = "multi", threshold = 0.5, mwg.sd, delta)
 ```
+In general we have the following proposal:
+$$
+\theta_{j, prop} \sim \N(\theta_{j, curr}, \sigma^2_{n,j})
+$$
+where $\theta_{j,prop}$ is the proposed (new) $j$-th parameter and $\theta_{j,curr}$ is the current (or old) $j$-th parameter.
+Besides, the adaptive standard deviation $\sigma_{n+1,j}$ at the $n+1$-th iteration for the $j$-th parameter is determined by
+$$
+\sigma_{n+1,j} = \exp\left(\log(\sigma_{n,j}) \pm \frac{\delta}{n} \right).
+$$
+Based on the above adaptive rule, we increase/decrease $\sigma_{n,j}$ at each step if the previous draw was accepted/rejected.
+The amount of adaptation will goes to $0$ as the number $n$ of iterations goes large.
+This kind of adaptation can preserve the MCMC stationary distribution.
+
 Then we can check the acceptance rate.
@@
 ```
-The acceptance rate is 94.6% which is close to the optimal 100%.
+The acceptance rate is 50.29% which is close to the optimal 45%.
 

### 2018-03-28 inst/doc/msde-pmcmc.R
keep using prior on Lambda1, Y2, fixed the analytic posterior via Kalman filter

@@ -37,10 +37,18 @@ names(fixed.params) <- bmod$param.names
 fixed.params["Lambda1"] <- FALSE
 
 ## ----prior---------------------------------------------------------------
-# prior on (Lambda1, Y_0)
+# prior on (Lambda1, Y2)
 hyper <- list(mu = c(0,0), Sigma = diag(2))
-names(hyper$mu) <- bmod$data.names
-dimnames(hyper$Sigma) <- rep(list(bmod$data.names), 2)
+names(hyper$mu) <- c("Lambda1", "Y2")
+dimnames(hyper$Sigma) <- rep(list(c("Lambda1", "Y2")), 2)
+
+## ----mcmc----------------------------------------------------------------
+# posterior sampling
+nsamples <- 1e5
+burn <- 1e3
+bpost <- sde.post(bmod, binit, hyper = hyper,
+                  fixed.params = fixed.params,
+                  nsamples = nsamples, burn = burn)
 
 ## ----L1-mcmc-------------------------------------------------------------
 L1.mcmc <- bpost$params[,"Lambda1"]
@@ -63,7 +71,7 @@ accept <- ppost$accept
 print(accept)
 
 ## ------------------------------------------------------------------------
-# posterior mean of theta
+# posterior samples
 L1.pmcmc <- ppost$params[ ,!fixed.params]
 
 ## ----pmcmc-plot, fig.width = 10, fig.height = 5, out.width = "90%"-------

### 2018-03-28 tests/dontrun/pmcmc-functions.R

@@ -1,49 +1,36 @@
-# helper functions for pmcmc
-
-# Approximation of log marginal likelihood log p(theta | y_T)
-#'@param x Vector of log weights (not incremental weights) at a given time
-#'@param nPart Number of particles
-#'@param thataPrior Prior of theta, by default it is 1
-#'@return log marginal approximated density log p_theta(theta | y_T)
-logY <- function(x, nPart, thetaPrior = 1) {
-  # check dimension compatibility
-  if(length(x) != nPart)
-    stop("Check the input vector, its length doesn't match the number of particles")
-  c   <- max(x)
-  ans <- log(sum(exp(x - c))/nPart) + c # to avoid possible overflow
-  ans <- ans * thetaPrior
-  return(ans)
-}
-
-# particle mcmc sampling
-#' @param model An \code{sde.model} object constructed with \code{\link{sde.make.model}}.
-#' @param init An \code{sde.init} object constructed with \code{\link{sde.init}}.
-#' @param theta0 Initial parameter values.
-#' @param fixed.params Logical vector of length \code{nparams} indicating which parameters are to be held fixed in the MCMC sampler.
-#' @param nsamples Number of particle MCMC iterations.
-#' @param npart Number of particles, fixed integer value.
-#' @param dT Scalar interobservation time.
-#' @param resample The type of particle resampling scheme to use. These are: multi(nomial), resid(ual), strat(ified), sys(tematic).
-#' @param threshold A scalar less than 1 to indicate the threshold for resampling. A negative number disables resampling.
-#' @param mwg.sd Standard deviation jump size for Metropolis-within-Gibbs on parameters and missing components of first SDE observation.
-#'
-#' @return A list of the following elements:
-#' \describe{
-#'   \item{\code{params}}{An \code{nsamples x nparams} matrix of posterior parameter draws.}
-#'   \item{\code{accept}}{A named list of acceptance rates for the various components of the MCMC sampler.}
-#'   \item{\code{init}}{The \code{sde.init} object which initialized the sampler.}
-#'   \item{\code{mwg.sd}}{A named vector of Metropolis-within-Gibbs standard devations used at the last posterior iteration.}
-#' }
-#'
-sde.pmcmc <- function(model, init, theta0, fixed.params, 
[…]

### 2018-03-28 tests/dontrun/test-pmcmc.R

@@ -1,14 +1,10 @@
 # multivariate OU model based on PMCMC
-
-# require(msde)
+require(msde)
 source("pmcmc-functions.R")
-
 # set a seed
 set.seed(123)
-
 # bivariate OU model
 bmod <- sde.examples("biou")
-
 # simulate some data
 # true parameter values
 Gamma0 <- .1 * crossprod(matrix(rnorm(4),2,2))
@@ -20,14 +16,12 @@ names(theta0) <- bmod$param.names
 # initial value
 Y0 <- rnorm(2)
 names(Y0) <- bmod$data.names
-
 # simulation
 dT <- runif(1, max = .1) # time step
 nObs <- 10
 bsim <- sde.sim(bmod, x0 = Y0, theta = theta0,
                 dt = dT, dt.sim = dT, nobs = nObs)
 YObs <- bsim$data
-
 # inference via MCMC
 binit <- sde.init(bmod, x = YObs, dt = dT, theta = theta0,
                   nvar.obs = 1) # second component is unobserved
@@ -35,11 +29,12 @@ binit <- sde.init(bmod, x = YObs, dt = dT, theta = theta0,
 fixed.params <- rep(TRUE, bmod$nparams)
 names(fixed.params) <- bmod$param.names
 fixed.params["Lambda1"] <- FALSE
-# prior on (Lambda1, Y_0)
-hyper <- list(mu = c(0,0), Sigma = diag(2))
-names(hyper$mu) <- bmod$data.names
-dimnames(hyper$Sigma) <- rep(list(bmod$data.names), 2)
-
+# prior on (Lambda1, Y2)
+hyper <- list(mu = c(0, 0), Sigma = diag(1, 2))
+#names(hyper$mu) <- c("Lambda1", bmod$data.names) # prior for (Lambda1, Y1, Y2)
+#dimnames(hyper$Sigma) <- rep(list(c("Lambda1", bmod$data.names)), 2)
+names(hyper$mu) <- c("Lambda1", "Y2")
+dimnames(hyper$Sigma) <- rep(list(c("Lambda1", "Y2")), 2)
 # posterior sampling
 nsamples <- 1e5
 burn <- 1e3
@@ -47,21 +42,20 @@ bpost <- sde.post(bmod, binit, hyper = hyper,
                   fixed.params = fixed.params,
                   nsamples = nsamples, burn = burn)
 L1.mcmc <- bpost$params[,"Lambda1"]
-
[…]

### 2018-04-28 inst/proj/sde.pmcmc.R
working on sde.pmcmc

@@ -1,0 +1,204 @@
+#' Particle MCMC sampler for the SDE posterior.
+#'
+#' Implementation of an independent-coordinate multivariate random-walk Metropolis sampler on the marginal posterior distribution of parameters and missing components of the first SDE observation (see Details).
+#'
+#' @param model An \code{sde.model} object constructed with \code{\link{sde.make.model}}.
+#' @param init An \code{sde.init} object constructed with \code{\link{sde.init}}.
+#' @param hyper The hyperparameters of the SDE prior. See \code{\link{sde.prior}}.
+#' @param nsamples Number of particle MCMC iterations.
+#' @param burn Integer number of burn-in samples, or fraction of \code{nsamples} to prepend as burn-in.
+#' @param rw.sd Standard deviation jump size for random walk proposal step size on parameters and missing components of first SDE observation.
+#' @param fixed.params Logical vector of length \code{nparams} indicating which parameters are to be held fixed in the MCMC sampler.
+#' @param last.miss.out Logical, whether to return the missing sde components of the last observation.
+#' @param npart Number of particles.
+#' @param resample The type of particle resampling scheme to use. These are: multi(nomial), resid(ual), strat(ified), sys(tematic).
+#' @param threshold A scalar less than 1 to indicate the threshold for resampling. A negative number disables resampling.
+#' @return A list of the following elements:
+#' \describe{
+#'   \item{\code{params}}{An \code{nsamples x nparams} matrix of posterior parameter draws.}
+#'   \item{\code{init}}{The \code{sde.init} object which initialized the sampler.}
+#'   \item{\code{rw.sd}}{A named vector of RW standard devations used at the last posterior iteration.}
+#'   \item{\code{accept}}{A named list of acceptance rates for the various components of the MCMC sampler.}
+#' }
+sde.pmcmc <- function(model, init, hyper,
+                      nsamples, burn, rw.sd = 1, fixed.params,
+                      last.miss.out = FALSE,
+                      npart, resample, threshold) {
+  # model constants
+  if(class(model) != "sde.model") {
[…]

### 2018-04-28 vignettes/sde.pmcmc.R

@@ -1,110 +1,0 @@
-# Particle MCMC sampling for SDE (currently not exported)
-# 
-# @param model An \code{sde.model} object constructed with \code{\link{sde.make.model}}.
-# @param init An \code{sde.init} object constructed with \code{\link{sde.init}}.
-# @param theta0 Initial parameter values.
-# @param fixed.params Logical vector of length \code{nparams} indicating which parameters are to be held fixed in the MCMC sampler.
-# @param hyper The hyperparameters of the SDE prior. For details, please see \code{\link{sde.prior}}.
-# @param nsamples Number of particle MCMC iterations.
-# @param npart Number of particles, fixed integer value.
-# @param dT Scalar interobservation time.
-# @param resample The type of particle resampling scheme to use. These are: multi(nomial), resid(ual), strat(ified), sys(tematic).
-# @param threshold A scalar less than 1 to indicate the threshold for resampling. A negative number disables resampling.
-# @param rw.sd Standard deviation jump size for random walk proposal step size on parameters and missing components of first SDE observation.
-# @param delta Tuning parameter for adaptive rw.sd. It has the same length as rw.sd
-#
-# @return A list of the following elements:
-# \describe{
-#   \item{\code{params}}{An \code{nsamples x nparams} matrix of posterior parameter draws.}
-#   \item{\code{accept}}{A named list of acceptance rates for the various components of the MCMC sampler.}
-#   \item{\code{init}}{The \code{sde.init} object which initialized the sampler.}
-#   \item{\code{rw.sd}}{A named vector of RW standard devations used at the last posterior iteration.}
-#   \item{\code{delta}}{A numeric value or a vector of tuning parameter(s) for rw.sd}
-# }
-#
-sde.pmcmc <- function(model, init, theta0, fixed.params, hyper, 
-                      nsamples, npart, dT,
-                      resample = "multi", threshold = 0.5, rw.sd = 1, delta = .618) {
-  YObs <- init$data
-  # iteration = 0, initialize the sde model for PMCMC
-  # get initial log-weights & log marginal likelihood via particle filtering
-  pf <- sde.pf(model, init, npart, resample, threshold, history = FALSE)
[…]

### 2021-12-09 tests/dontrun/Rcpp_templates.Rmd
implemented OOP via RC, passes Rdevel CMD check --use-valgrind

---
title: "C++ class templates with Rcpp and R6"
author: "Martin Lysy"
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
cran_link <- function(pkg) {
  paste0("[**", pkg, "**](https://CRAN.R-project.org/package=", pkg, ")")
}
```

# Motivation

Consider the following paradigm for writing methods for parameter inference that can be applied to any user-defined member of a family of models.  This is achieved using a templated class `Inference<Model>` of which the template parameter `Model` defines model-specific versions of methods which `Inference` might use (e.g., `Model.loglik()`).  Within the proposed paradigm we'll make the following assumption:

> The only public members of the `Inference<Model>` class (that get exported to R) all have the same signature, including the class constructor, and all have arguments and return values than can be wrapped by `r cran_link("Rcpp")`.

So for example:
```cpp
/// User-defined model.
class IIDNormal {
 public:
  double loglik(const NumericVector& theta, const NumericVector& x) {
	double ll = 0.0;
	for(int ii=0; ii<x.length(); ii++) {
	  ll += R:dnorm(x(ii), theta(0), theta(1), true);
	}
	return ll;
  }
};

/// Generic methods for inference.
template <class Model>
class Inference {
 private:
  Model model_;
 public:
  // Calculate mle based on x and model.loglik(theta, x)
  void mle(NumericVector& theta_hat, const NumericVector);
};

// C++ API
iidnorm_fit = Inference<IIDNormal>; // create inference object
iidnorm_fit.mle(theta_hat, x); // calculate MLE
```

The objective is to create an R wrapper to `Inference<IIDNormal>` and its public methods.  It is proposed to do this using **Rcpp** and `r cran_link("R6")` classes as follows.

# **Rcpp** Wrapper

We could use **Rcpp** [modules](https://dirk.eddelbuettel.com/code/rcpp/Rcpp-modules.pdf) for this.  However, there are a few caveats to this approach:

- The R wrapper is an RC class, not R6, but the latter is much more lightweight, flexible, and easy to use.

- The R wrapper provides methods which accept arguments positionally and with zero error checking.  This is in contrast to how **Rcpp** would export the following function:

	```cpp
	//[[Rcpp::export]]
	double foo(double x, NumericVector y)
	```

    That is, the R call `foo(y = 1:5, x = 3)` would work as expected and `foo(1:5, x)` would give an error, whereas `obj$foo()` as a class method exported by **Rcpp** modules would in both cases crash the R session.

Bearing this in mind, the approach advocated here is to create something like `Inference_IIDNormal_Exports.cpp` which wraps the class functionality in a rudimentary but transparent way.  Since we don't want to do this manually for each `Model`, we can use the [Mustache](https://mustache.github.io/) templating logic implemented in the R package `r cran_link("whisker")`.  So for example the template `cpp` file would look like this:

```
/// @file Inference_{{Model}}_Exports.cpp
///
/// Rcpp wrapper for the methods of Inference<{{Model}}>

#include {{Model_file}} // {{Model}} class definition
#include Inference.h // Inference class definition

//[[Rcpp::export]]

```

# R6 Wrapper

On the R side, we have an R6 class called `Inference_IIDNormal`.

### 2021-12-09 R/msde-package.R

@@ -1,7 +1,3 @@
-#' Simulation and inference for multivariate stochastic differential equations.
-#'
-#' @docType package
-#' @name msde
 #' @useDynLib msde, .registration = TRUE
 #' @details See package vignettes; \code{vignette("msde-quicktut")} for a tutorial and \code{vignette("msde-exmodels")} for several example models.
 #' @examples
@@ -45,4 +41,4 @@
 #'        main = parse(text = hmod$param.names[ii]), xlab = "")
 #' }
 #' }
-NULL
+"_PACKAGE"

### 2021-12-09 R/sde.drift.R

@@ -43,10 +43,10 @@ sde.drift <- function(model, x, theta) {
   if(!all(.is.valid.data(model, x, theta, single.x, single.theta, nreps))) {
     stop("x contains invalid sde data.")
   }
-  dr <- .sde_Drift(sdeptr = model$ptr, xIn = as.double(x),
-                   thetaIn = as.double(theta),
-                   singleX = as.logical(single.x),
-                   singleTheta = as.logical(single.theta),
-                   nReps = as.integer(nreps))
+  dr <- model$cobj$Drift(xIn = as.double(x),
+                         thetaIn = as.double(theta),
+                         singleX = as.logical(single.x),
+                         singleTheta = as.logical(single.theta),
+                         nReps = as.integer(nreps))
   matrix(dr, nrow = nreps, ncol = model$ndims, byrow = TRUE)
 }

### 2021-12-09 R/sde.examples.R

@@ -30,32 +30,32 @@ sde.examples <- function(model = c("hest", "pgnet", "lotvol", "biou", "eou"),
     ModelFile <- file.path(.msde_include_path, "hestModel.h")
     param.names <- c("alpha", "gamma", "beta", "sigma", "rho")
     data.names <- c("X", "Z")
-    sptr <- .hest_MakeModel()
+    cobj <- new(msde_hestModel)
   } else if(model == "pgnet") {
     ModelFile <- file.path(.msde_include_path, "pgnetModel.h")
     param.names <- paste0("theta", 1:8)
     data.names <- c("R", "P", "Q", "D")
-    sptr <- .pgnet_MakeModel()
+    cobj <- new(msde_pgnetModel)
   } else if(model == "biou") {
     ModelFile <- file.path(.msde_include_path, "biouModel.h")
     param.names <- c("Gamma11", "Gamma21", "Gamma12", "Gamma22",
                      "Lambda1", "Lambda2",
                      "Psi11", "Psi12", "Psi22")
     data.names <- c("Y1","Y2")
-    sptr <- .biou_MakeModel()
+    cobj <- new(msde_biouModel)
   } else if(model == "lotvol") {
     ModelFile <- file.path(.msde_include_path, "lotvolModel.h")
     param.names <- c("alpha", "beta", "gamma")
     data.names <- c("H", "L")
-    sptr <- .lotvol_MakeModel()
+    cobj <- new(msde_lotvolModel)
   } else if(model == "eou") {
     ModelFile <- file.path(.msde_include_path, "eouModel.h")
     param.names <- c("alpha", "gamma", "eta", "sigma", "rho")
     data.names <- c("X", "V")
-    sptr <- .eou_MakeModel()
+    cobj <- new(msde_eouModel)
   }
   if(!file.only) {
-    sde.model <- list(ptr = sptr,
+    sde.model <- list(cobj = cobj,
                       ndims = length(data.names), nparams = length(param.names),
                       data.names = data.names, param.names = param.names,
                       hyper.check = mvn.hyper.check, omp = FALSE)

### 2021-12-09 R/sde.is.valid.R

@@ -44,11 +44,11 @@ sde.valid.data <- function(model, x, theta) {
     stop("x and theta have incompatible dimensions.")
   }
   nreps <- max(nreps)
-  .sde_isData(sdeptr = model$ptr, xIn = as.double(x),
-              thetaIn = as.double(theta),
-              singleX = as.logical(single.x),
-              singleTheta = as.logical(single.theta),
-              nReps = as.integer(nreps))
+  model$cobj$isData(xIn = as.double(x),
+                    thetaIn = as.double(theta),
+                    singleX = as.logical(single.x),
+                    singleTheta = as.logical(single.theta),
+                    nReps = as.integer(nreps))
 }
 
 #' @rdname sde.valid
@@ -60,23 +60,23 @@ sde.valid.params <- function(model, theta) {
   theta <- .format.params(theta, model$param.names)
   # check singles and compatible x and theta
   nreps <- ncol(theta)
-  .sde_isParams(sdeptr = model$ptr, thetaIn = as.double(theta),
-                nReps = as.integer(nreps))
+  model$cobj$isParams(thetaIn = as.double(theta),
+                      nReps = as.integer(nreps))
 }
 
 #--- internal versions: no argument checking/formatting -------------------
 
 .is.valid.data <- function(model, x, theta, single.x, single.theta,
                            nreps) {
-  .sde_isData(sdeptr = model$ptr, xIn = as.double(x),
-              thetaIn = as.double(theta),
-              singleX = as.logical(single.x),
-              singleTheta = as.logical(single.theta),
-              nReps = as.integer(nreps))
+  model$cobj$isData(xIn = as.double(x),
+                    thetaIn = as.double(theta),
+                    singleX = as.logical(single.x),
+                    singleTheta = as.logical(single.theta),
+                    nReps = as.integer(nreps))
 }
 
 .is.valid.params <- function(model, theta, single.theta, nreps) {
-  rep(.sde_isParams(sdeptr = model$ptr, thetaIn = as.double(theta),
-                    nReps = as.integer(ifelse(single.theta, 1, nreps))),
+  rep(model$cobj$isParams(thetaIn = as.double(theta),
+                          nReps = as.integer(ifelse(single.theta, 1, nreps))),
       times = ifelse(single.theta, nreps, 1))
 }

### 2021-12-09 R/sde.loglik.R

@@ -71,12 +71,12 @@ sde.loglik <- function(model, x, dt, theta, ncores = 1) {
     stop("x contains invalid sde data.")
   }
   # compute
-  .sde_Loglik(sdeptr = model$ptr,
-              xIn = as.double(x), dTIn = as.double(dt),
-              thetaIn = as.double(theta),
-              nComp = as.integer(ncomp),
-              nReps = as.integer(nreps),
-              singleX = as.logical(single.x),
-              singleTheta = as.logical(single.theta),
-              nCores = as.integer(ncores))
+  model$cobj$Loglik(xIn = as.double(x),
+                    dTIn = as.double(dt),
+                    thetaIn = as.double(theta),
+                    nComp = as.integer(ncomp),
+                    nReps = as.integer(nreps),
+                    singleX = as.logical(single.x),
+                    singleTheta = as.logical(single.theta),
+                    nCores = as.integer(ncores))
 }

### 2021-12-09 R/sde.post.R

@@ -148,29 +148,28 @@ sde.post <- function(model, init, hyper,
 #  if(debug) browser()
   tm <- chrono()
   # compute
-  ans <- .sde_Post(sdeptr = model$ptr,
-                   initParams = as.double(init.params),
-                   initData = as.double(t(init.data)),
-                   dT = as.double(dt),
-                   nDimsPerObs = as.integer(par.index),
-                   fixedParams = as.logical(fixed.params),
-                   nSamples = as.integer(nsamples),
-                   burn = as.integer(burn),
-                   nParamsOut = as.integer(nparams.out),
-                   nDataOut = as.integer(ndata.out),
-                   dataOutSmp = as.integer(data.out$isamples-1),
-                   dataOutComp = as.integer(data.out$icomp-1),
-                   dataOutDims = as.integer(data.out$idims-1),
-                   updateParams = as.double(update.params),
-                   updateData = as.double(update.data),
-                   priorArgs = prior,
-                   tunePar = tune.par,
-                   updateLogLik = as.integer(loglik.out),
-                   nLogLikOut = as.integer(ifelse(loglik.out, nsamples, 1)),
-                   updateLastMiss = as.integer(last.miss.out),
-                   nLastMissOut = as.integer(nlast.miss.out),
-                   nCores = as.integer(ncores),
-                   displayProgress = as.logical(verbose))
+  ans <- model$cobj$Post(initParams = as.double(init.params),
+                         initData = as.double(t(init.data)),
+                         dT = as.double(dt),
+                         nDimsPerObs = as.integer(par.index),
+                         fixedParams = as.logical(fixed.params),
+                         nSamples = as.integer(nsamples),
+                         burn = as.integer(burn),
+                         nParamsOut = as.integer(nparams.out),
+                         nDataOut = as.integer(ndata.out),
+                         dataOutSmp = as.integer(data.out$isamples-1),
+                         dataOutComp = as.integer(data.out$icomp-1),
+                         dataOutDims = as.integer(data.out$idims-1),
+                         updateParams = as.double(update.params),
+                         updateData = as.double(update.data),
+                         priorArgs = prior,
+                         tunePar = tune.par,
+                         updateLogLik = as.integer(loglik.out),
+                         nLogLikOut = as.integer(ifelse(loglik.out, nsamples, 1)),
+                         updateLastMiss = as.integer(last.miss.out),
[…]

### 2021-12-09 R/sde.prior.R

@@ -53,10 +53,11 @@ sde.prior <- function(model, theta, x, hyper) {
     stop("model$hyper.check must convert hyper to a list with NULL or numeric-vector elements.")
   }
   # compute
-  ans <- .sde_Prior(sdeptr = model$ptr,
-                    thetaIn = as.double(theta), xIn = as.double(x),
-                    singleTheta = as.logical(single.theta),
-                    singleX = as.logical(single.x),
-                    nReps = as.integer(nreps), phiIn = phi)
+  ans <- model$cobj$Prior(thetaIn = as.double(theta),
+                          xIn = as.double(x),
+                          singleTheta = as.logical(single.theta),
+                          singleX = as.logical(single.x),
+                          nReps = as.integer(nreps),
+                          phiIn = phi)
   ans
 }

### 2021-12-09 R/zzz.R

@@ -7,4 +7,9 @@
   assign(".msde_tools_path",
          file.path(libname, pkgname, "include", "R"),
          envir = parent.env(environment()))
+  loadModule("class_msde_hestModel", TRUE)
+  loadModule("class_msde_lotvolModel", TRUE)
+  loadModule("class_msde_biouModel", TRUE)
+  loadModule("class_msde_pgnetModel", TRUE)
+  loadModule("class_msde_eouModel", TRUE)
 }

### 2021-12-09 inst/include/R/sdeMakeModel.cpp

@@ -3,9 +3,7 @@
 // (rest of include directory is not, except rngUtils.h)
 
 //[[Rcpp::depends("msde")]]
-//[[Rcpp::depends("RcppArmadillo")]]
 //[[Rcpp::depends("RcppProgress")]]
-#include <RcppArmadillo.h>
 using namespace Rcpp;
 #include <sdeInterface.h>
 #include "sdeModel.h"

### 2021-12-09 inst/include/sdeRUtils.h

@@ -1,9 +1,7 @@
 #ifndef sdeRUtils_h
 #define sdeRUtils_h
 
-// #include <Rcpp.h>
-//[[Rcpp::depends("RcppArmadillo")]]
-#include <RcppArmadillo.h>
+#include <Rcpp.h>
 typedef Rcpp::LogicalVector Logical;
 typedef Rcpp::NumericVector Numeric;
 typedef Rcpp::IntegerVector Integer;
@@ -11,7 +9,7 @@ typedef Rcpp::List List;
 //using namespace Rcpp;
 #include "sdeUtils.h"
 #include "sdeLogLik.h"
-#include "sdeInterface.h"
+#include "sdeRobj.h"
 
 template <class sMod, class sPi>
   inline int sdeRobj<sMod, sPi>::get_nDims() {

### 2021-12-09 inst/include/sdeRobj.h

@@ -1,0 +1,165 @@
+///
+///
+/// Interface considerations for the implementation.
+///
+/// 1. `sde.make.model` should create an `Xptr` instead of returning R/C++ entrypoints for all functions.  Therefore, it should trigger the construction of an `sdeInterface` object of which the members interface between R/C++ for the necessary functions: `drift()`, `diff()`, `sim()`, `post()`, etc.
+/// 2. must be able to hold different models in one R session.  This means that Xptr should be a templated class.  That is, each class/function making use of `sdeModel` and `sdePrior` should accept templates of these.
+/// 3. should have pre-compiled versions of some models.  For this, the src of the package should have pointers to different models in its shared object...
+///
+/// @section C++ API
+/// 
+/// ```
+/// // Abstract base, template derived class
+/// sdeCobj : sdeRobj<sdeModel, sdePrior>
+/// 
+/// // R constructor:
+/// sdeCobj *sde = new sdeRobj<sdeModel, sdePrior>;
+/// XPtr<sdeRobj> sdeptr(sde, true);
+///
+/// return sdeptr;
+///
+/// @section  R Methods
+/// 
+/// - `sde.drift()` and `sde.diff()`.
+/// - `sde.loglik()`.
+/// - `sde.prior()`.
+/// - `sde.valid.params()` and `sde.valid.data()`.
+/// - `sde.sim()`.
+/// - `sde.post()`.
+
+#ifndef sdeRobj_h
+#define sdeRobj_h
+
+#include <Rcpp.h>
+typedef Rcpp::LogicalVector Logical;
+typedef Rcpp::NumericVector Numeric;
+typedef Rcpp::IntegerVector Integer;
+typedef Rcpp::NumericMatrix NumericMatrix;
+typedef Rcpp::List List;
+//using namespace Rcpp;
+//#include "sdeLogLik.h"
+//#include "sdeMCMC.h"
+//#include "mcmcUtils.h"
+
+// --- helper functions --------------------------------------------------------
+
+// R -> C++ logical vector conversion.
+inline void convert_Logical(bool *out, Logical in) {
+  for(int ii=0; ii<in.length(); ii++) {
+    if(Logical::is_na(in[ii])) {
+      Rprintf("convert_Logical: NA detected.\n");
+    }
+    out[ii] = (in[ii] != 0);
+  }
[…]

### 2021-12-09 inst/include/templates/sde.model.R

@@ -1,0 +1,112 @@
+#' R6 class wrapping `sdeRobj<{{sdeModel}}, {{sdePrior}}>`.
+#'
+#' @details This class is intended to be hidden, i.e., wouldn't be exported from a package.  Rather, it is the instantiated object which would potentially be exported, since it would be called with the `sde.{method}` generics provided by **msde**.
+#'
+#' That being said, the package would probably want to export the class constructor, so as to instantiate new models...
+#' @noRd
+{{R6ClassName}} <- R6::R6Class(
+
+  classname = "sde.model",
+
+  private = list(
+    .ptr = NULL,
+    .ndims = NULL,
+    .nparams = NULL,
+    .data.names = NULL,
+    .param.names = NULL,
+    .omp = NULL
+  ),
+
+  active = list(
+
+    #' @field ptr Pointer to C++ `sdeRobj<{{sdeModel}}, {{sdePrior}}>`.
+    ptr = function(value) {
+      if(missing(value)) {
+        private$.ptr
+      } else {
+        stop("$ptr is read-only.", call. = FALSE)
+      }
+    },
+
+    #' @field ndim Number of SDE components.
+    ndim = function(value) {
+      if(missing(value)) {
+        private$.ndim
+      } else {
+        stop("$ndim is read-only.", call. = FALSE)
+      }
+    },
+
+    #' @field nparams Number of model parameters.
+    nparams = function(value) {
+      if(missing(value)) {
+        private$.nparams
+      } else {
+        stop("$nparams is read-only.", call. = FALSE)
+      }
+    },
+
+    #' @field data.names Names of the SDE components.
+    data.names = function(value) {
+      if(missing(value)) {
+        private$.data.names
+      } else {
+        if(length(value) != self$ndims) {
+          stop("$data.names has wrong length.")
+        } else {
+          private$.data.names <- value
+        }
+      }
+    },
+
+    #' @field param.names Names of the model parameters.
+    param.names = function(value) {
+      if(missing(value)) {
+        private$.param.names
+      } else {
+        if(length(value) != self$nparams) {
+          stop("$param.names has wrong length.")
+        } else {
+          private$.param.names <- value
+        }
+      }
+    },
+
+    #' @field omp A logical flag for whether or not the model was compiled for multicore functionality with \code{OpenMP}.
+    omp = function(value) {
[…]

### 2021-12-09 src/old/sdeExports.cpp

@@ -1,8 +1,8 @@
 // generic R wrappers to exported members of instantiated Cobj
 
 // #include <Rcpp.h>
-//[[Rcpp::depends("RcppArmadillo")]]
-#include <RcppArmadillo.h>
+// //[[Rcpp::depends("RcppArmadillo")]]
+// #include <RcppArmadillo.h>
 using namespace Rcpp;
 #include "sdeInterface.h"
 

### 2021-12-09 tests/dontrun/test-interface.R

@@ -1,5 +1,41 @@
 #--- msde interface ------------------------------------------------------------
 
+require(msde)
+require(whisker)
+require(Rcpp)
+
+msde.path <- "/Users/mlysy/Documents/proj/msde/msdeHeaders"
+
+template.data <- list(
+  ExportFile = "msde_hestModel_Exports.cpp",
+  ## ModelFile = system.file("include", "hestModel.h", package = "msde"),
+  ## PriorFile = system.file("include", "mvnPrior.h", package = "msde"),
+  ModelFile = "hestModel.h",
+  PriorFile = "mvnPrior.h",
+  sdeModel = "sdeModel",
+  sdePrior = "sdePrior",
+  ## cppClassName = "msde_hestModel",
+  ModuleName = "msde_hestModel",
+  RClassName = "msde_hestModel" # don't start name with `.`
+)
+
+out <- whisker.render(
+  template = readLines(file.path(msde.path,
+                                 "inst", "include", "templates",
+                                 "sdeModule.cpp")),
+  data = template.data
+)
+
+cat(out, sep = "\n")
+
+sourceCpp(code = out, verbose = TRUE)
+
+obj <- eval(parse(text = paste0("new(", "msde_hestModel", ")")))
+
+sde.make.model(ModelFile = file.path(msde.path, "inst", "include", "hestModel.h"))
+
+#--- older ---------------------------------------------------------------------
+
 devtools::document()
 devtools::install()
 

### 2021-12-17 tests/dontrun/msde-pmcmc.Rmd
added pkgdown

--- 2018-03-22 inst/doc/msde-pmcmc.Rmd
@@
 title: "Particle MCMC for Multivariate Stochastic Differential Equations with **`msde`**"
-author: "Martin Lysy, Feiyu Zhu"
+author: "`r readLines('authors.txt')`"
 date: "`r Sys.Date()`"
@@
 ```
-Then we set the prior on the unknown parameter(s) and observations.
+Then we set the prior on the unknown parameter $\Lambda_1$ and the latent observation $Y_2$. The other elements are given a Lebesgue prior.
+$$
+\Lambda_1, Y_2 \overset{iid}{\sim} \N(0,1), \quad \pi(Y_1, \Gamma, \Lambda_2, \Phi | \Lambda_1, Y_2) \propto 1.
+$$ 
 ```{r prior}
-# prior on (Lambda1, Y_0)
+# prior on (Lambda1, Y2)
 hyper <- list(mu = c(0,0), Sigma = diag(2))
-names(hyper$mu) <- bmod$data.names
-dimnames(hyper$Sigma) <- rep(list(bmod$data.names), 2)
+names(hyper$mu) <- c("Lambda1", "Y2")
+dimnames(hyper$Sigma) <- rep(list(c("Lambda1", "Y2")), 2)
 ```
 We set the total number of posterior samples as `100,000` with the burn-in period as `1,000`. The posterior sampling via MCMC is given as follows:
-```{r mcmc, cache = TRUE}
+```{r mcmc}
 # posterior sampling
@@
   lambda[1] <- l1
-  mou.loglik(X = YObs, dt = dT, nvar.obs = 1,
+  tmp <- mou.loglik(X = YObs, dt = dT, nvar.obs = 1,
                   Gamma = Gamma0, Lambda = lambda, Phi = Phi0,
-             mu0 = hyper$mu, Sigma0 = hyper$Sigma)
+                  mu0 = c(0, 0), Sigma0 = diag(2))
+  tmp <- tmp + dnorm(l1, 0, 1, log = TRUE) # add the prior log density of Lambda1
+  return(tmp)
 })
@@
 ```{r source, echo = FALSE}
-source("sde.pmcmc.R")
+source(system.file("proj", "sde.pmcmc.R", package = "msde"))
 ```
@@
 npart <- 100
-mwg.sd <- 1
-delta <- .618
+rw.sd <- 1
+delta <- .7
 ppost <- sde.pmcmc(model = bmod, binit, theta0, fixed.params, hyper,
                    nsamples, npart, dT,
-                   resample = "multi", threshold = 0.5, mwg.sd, delta)
+                   resample = "multi", threshold = 0.5, rw.sd, delta)
 ```
@@
 ```{r}
-# posterior mean of theta
+# posterior samples
 L1.pmcmc <- ppost$params[ ,!fixed.params]
